Olfactory basis for essential amino acid perception during foraging in Caenorhabditis elegans

Version 1

Abstract

Abstract

Acquisition of essential nutrients through diet is crucial for the survival of animals. Dietary odors might enable foraging in animals for nutrient-rich diets. We asked if Caenorhabditis elegans, a bacterivorous nematode, uses olfactory cues to forage for essential amino acid-rich (EAA) diets. Using the native microbiome of C. elegans, we show that worms rely on olfaction to select leucine (EAA)-enriched bacteria. Using gas chromatography, we find that leucine-enriched bacteria produce isoamyl alcohol (IAA) odor in the highest abundance. Prior adaptation of worms to IAA diminishes the diet preference of worms. Several wild isolates of C. elegans display robust responses to IAA emphasizing its ecological relevance. We find that foraging for a leucine-enriched diet is mediated via the AWC olfactory neurons. Finally, we identify SRD-12, in AWC neurons, as a receptor for IAA and a mediator of dietary decisions in worms. Our study identifies a receptor-ligand module underpinning foraging behavior in C. elegans.

Introduction

Amino acids are central to metabolism, not only as building blocks for proteins but also as signaling molecules. During the course of evolution, animals have lost the ability to synthesize half of the proteogenic amino acids, making dietary intake of such amino acids indispensable 1. Leucine (Leu), isoleucine (Ile), valine (Val), histidine (His), lysine (Lys), tryptophan (Trp), phenylalanine (Phe), methionine (Met), threonine (Thr), and arginine (Arg) constitute essential amino acids (EAAs) 2. Several studies have investigated the mechanism of amino acid surveillance by cells and tissues. Leucine, glutamine, and arginine have been shown to activate the mTOR pathway and control protein translation and cell growth 3–9. Altered EAA levels influence the longevity, behavior, and health of animals 10–12. Increased levels of circulating branched-chain amino acids (BCAAs) are associated with an increased risk of obesity and diabetes in humans 13,14. Restricted intake of tryptophan and methionine extends lifespan in rats, while a tryptophan-deficient diet reduces the weight of organs 15,16. Methionine restriction is reported to extend the lifespan of Drosophila 17. Central administration of a BCAA, leucine, reduces food intake in rats 18. Extended lifespan upon diet restriction in silkworms is linked to enhanced utilization of EAAs 19. However, it is not clear if animals have a way to modulate EAA-specific dietary intake. Do animals use specific sensing mechanisms to find an EAA-enriched diet? Several studies on feeding behaviors suggest that both invertebrates and vertebrates select nutrient-rich diets. For instance, rats prefer a protein-rich diet over a carbohydrate-rich diet 20. Many animals display mechanisms to choose for or reject food with specific EAAs. Sparrows consume combination of different diets to get adequate levels of various EAAs, while gastropods reject diets deficient in methionine 21–23. When maintained on an EAA-restricted diet, flies compensate by increasing feeding 24. Kittens prefer diets rich in methionine and threonine, while rats select for a leucine-enriched diet 25. Despite evidence that organisms seek EAA, specific molecular mechanisms for search behavior are not well understood.

Chemoperception, taste and olfaction, allow animals to quickly assess their diet utilizing G-protein coupled receptors (GPCRs). EAAs evoke taste perceptions such as sweet (Thr, His, Leu, Phe, Trp) and bitter (Arg, Ile, Lys, Met), while non-EAAs can elicit umami (Glu) and sour (Asp) taste 26–28. Amino acids that elicit sweet responses in humans, like threonine, are found to be attractive in rodents 29,30. GPCRs are implicated in sensing different classes of amino acids and regulate foraging behavior in animals 31–35. Metabotropic glutamate receptors (mGlu) sense glutamate, while the extracellular calcium-sensing receptor senses aliphatic, aromatic, and polar amino acids 34,36. Invertebrates like Caenorhabditis elegans and Drosophila melanogaster also possess GPCRs-based amino acid sensing mechanisms 37. C. elegans senses glutamate using mGlu type receptors, Mgl-1/2/3, while D. melanogaster utilizes a cationic amino-acid transporter, Dm slif, to sense arginine 38,39. The goldfish 5.24, an odorant receptor, responds to arginine, while its human homolog, GPRC6A, responds to basic as well as aliphatic amino acids 40,41. Olfactory GPCRs are known to sense various odors, including alcohol, esters, ketones, etc. and many of these are products of EAA catabolism. It is not known if animals rely on EAA-derived odors to find an EAA-rich diets.

The presence of an elaborate chemosensory system in soil-dwelling animals suggests the usage of olfactory cues for foraging. C. elegans, a bacterivorous nematode, relies heavily on its simple but effective chemosensory system composed of 302 neurons for foraging, although specific cues regulating this phenomenon are unknown 42,43. Olfactory neurons of C. elegans express as many as 335 GPCRs 44, however, the odors stimulating them remain largely unknown. Like other invertebrates, C. elegans requires ten EAAs, of these, leucine, tryptophan, valine, arginine, and lysine extend their lifespan when supplemented at low concentration 45. Does C. elegans rely on olfactory cues to sense diets with high(er) levels of EAAs? It is conceivable since some bacteria possess biosynthetic pathways to convert amino acids into odors, such as dimethyl sulfide, indole, isoamyl acetate, phenylethyl alcohol, and benzyl alcohol 46. C. elegans can sense over 120 odors representing extensive structural diversity 47. Do worms utilize some of them as foraging signals to find nutrient-enriched bacteria?.

We hypothesized that certain odors produced by the microbiome of C. elegans are olfactory signals for EAA-rich bacteria and drive worms’ foraging behavior. To test this hypothesis, we analyzed the preference of worms for natural microbiota species supplemented with individual EAAs. We found that odors drive worms’ feeding preference in favor of leucine supplemented diets. By carefully analyzing the odor bouquet produced by natural microbiota upon leucine supplementation, we identified isoamyl alcohol (IAA) as the most abundant odor in preferred diets. We show that IAA is produced from leucine using the Ehrlich degradation pathway in bacteria. IAA is an ecologically relevant odor and regulates the preference of worms for a leucine supplemented diet. Finally, we identified SRD-12, a GPCR expressed in AWC neurons, as the receptor for IAA. Taken together, SRD-12 regulates the dietary preference of worms to IAA-producing bacteria and thereby mediates the foraging behavior of C. elegans to leucine-enriched diets. Thus, IAA produced by bacteria is a diet quality code for leucine-enriched bacteria.

Results

Microbial odors drive the preference of C. elegans for leucine-enriched diets

To understand if C. elegans used olfactory signals to distinguish between standard and EAA-enriched diets, we performed ‘odor-only’ diet preference assays using natural microbiota of C. elegans, CeMbio, and laboratory food, Escherichia coli OP50. CeMbio is a collection of 12 bacteria that best reflects the metabolic complexity of the microbes in the natural habitat of worms (see Table S1). N2 or wild type (WT) worms were allowed to choose between odors from each strain grown on separate sections, EAA supplemented (+ EAA) or unsupplemented (-EAA), of a tripartite plate (schematic in Figure 1A). Of the ten EAAs tested, WT worms showed a preference only for bacteria on leucine supplemented (+ LEU) over unsupplemented (-LEU) for three microbial strains, namely, Enterobacter hormaechei (CEent1), Lelliottia amnigena (JUb66), and Sphingobacterium multivorum (BIGb0170) (Figure 1B). However, worms were not able to distinguish between supplemented and unsupplemented bacteria for the remaining nine EAAs (Figures 1C, 1D and S1A-S1G). We found that worms did not respond to leucine at three different concentrations in a chemotaxis assay, suggesting that worms are incapable of sensing leucine directly (Figure S1H). These results suggested that worms can forage for diets enriched in a specific EAA, leucine, using odors produced by these bacteria.

C. elegans relies on odors to select leucine-enriched bacteria.

(A) Schematic representation of ‘odor-only’ diet preference assay.

Preference index (PI) of WT worms in an ‘odor-only’ diet preference assay (indicated as ) for individual diet supplemented with (B) 5 mM leucine (+ LEU), (C) 5 mM isoleucine (+ ILE), and (D) 5 mM valine (+ VAL). Error bars indicate SEM (n≥15). Also see Figure S1.

We next asked whether worms prefer their native microbiome over laboratory food E. coli OP50. We performed ‘odor-only’ diet preference assays to test if worms can distinguish between CeMbio and E. coli OP50 using odors (schematic in Figure S2A). We found that the preference indices (PI) of worms for CEent1, JUb66, and BIGb0170 were positive, making them the preferred diets, while neutral or negative for other strains such as Pseudomonas lurida MYb11 (Figure S2B). These findings suggested that worms not only rely on odors to choose between two bacteria, but also to find leucine-enriched bacteria.

Isoamyl alcohol odor is a signature for a leucine-enriched diet

We hypothesized that bacterial odor(s) enriched upon leucine supplementation are likely to be drivers of foraging behavior in worms. We identified all the odors in the headspace of individual bacteria using solid-phase microextraction followed by gas chromatography-mass spectrometry (GC-MS/MS). By examining levels of individual odors in leucine supplemented and unsupplemented bacteria, we found that several odors were present in the headspace of bacteria (Figures 2A, 2B, and S3A-S3D). However, isoamyl alcohol (IAA) was the most abundant and the only shared odor in the worms’ preferred diets, CEent1, JUb66, and BIGb0170 (Figures 2A-2C and S3A-S3D). We found that under standard conditions, IAA constituted 40-70% of the headspace of preferred bacteria and remarkably increased up to 90% upon leucine supplementation (Figure 2C). We also quantified leucine-dependent increase in IAA levels in the headspace of preferred bacteria (Figure S3E, see methods). We found that abundance of IAA significantly increased in CEent1, JUb66, and BIGb0170 by 1.2 - 4 folds upon leucine supplementation (Figures 2D, S3F and S3G).

Leucine supplementation boosts isoamyl alcohol levels via Ehrlich degradation pathway.

GC-MS/MS profile of odors produced by CEent1 under (A) leucine supplemented and (B) leucine unsupplemented conditions. Unmarked peaks represent masses contributed by fiber or media alone (n≥3).

(C) Heat map representing the relative abundance of various odors in the headspace of CEent1, JUb66, and BIGb0170 with and without leucine supplementation.

(D) Absolute abundance of IAA (in moles) produced by CEent1 under leucine supplemented and unsupplemented conditions. * P ≤ 0.05 as determined by a two-tailed unpaired t-test.

(E) Schematic of Ehrlich degradation pathway.

(F) Fold change of transcript levels of ilvE under leucine supplemented over unsupplemented conditions for CEent1.

Error bars indicate SEM (n≥3).

Microbes can produce short-chain alcohols from carbon sources as well as branched-chain amino acids 46. Leucine can be catabolized to IAA via the Ehrlich degradation pathway in three enzymatic steps. The first step is catalyzed by IlvE, a transaminase, which converts leucine to α-Ketoisocaproate (schematic in Figure 2E). The ilvE gene is present in the genome of CEent1, JUb66, and BIGb0170 48. We predicted that leucine supplementation would result in the upregulation of ilvE in a substrate-dependent manner. Using qRT-PCR, we found that transcript levels of ilvE were 2-fold higher in leucine supplemented CEent1 over unsupplemented (Figure 2F). This indicated that the Ehrlich degradation pathway is functional in IAA-producing bacteria and further stimulated in response to leucine. Altogether, these findings show that IAA produced via Ehrlich degradation is a signature for a leucine-enriched bacteria.

AWC odor sensory neurons facilitate the diet preference of C. elegans for leucine-enriched diets

Worms use two pairs of odor sensory neurons, AWA and AWC, to sense attractive olfactory cues 47,49,50. To identify the odor sensory neurons that mediate preference for leucine-enriched diets, we studied the response of odr-7 worms, which do not have functional AWA neurons, and AWC ablation (-) worms 51,52. We found that the preference of worms for leucine supplemented diets was dramatically lost in AWC(-) worms but was retained in odr-7 worms (Figures 3A-3C). This suggested that AWC neurons play a crucial role in sensing odors from leucine-enriched diets. We also tested the role of these neurons in mediating diet preference between preferred CeMbio bacteria and E. coli OP50. We found that AWC(-) worms showed reduced preference for CEent1, JUb66 and BIGb0170 over E. coli, while odr-7 mutants had slightly diminished preference for CEent1 and BIGb0170 over E. coli (Figures S4A-S4C). This suggested AWC, predominantly, and AWA odor sensory neurons contribute to the diet preference of worms for specific microbes in their natural habitat (Figure S4D). The finding that diet preference entails two pairs of neurons suggests that multiple odors constitute foraging signal enabling worms to find diet bacteria.

Odors sensed through AWC neurons mediate the diet preference of C. elegans.

Preference index (PI) of WT, odr-7, and AWC(-) worms for (A) CEent1, (B) JUb66, and (C) BIGb0170 supplemented with leucine over unsupplemented conditions. symbol indicates ‘odor-only’ preference assays. Chemotaxis index (CI) of WT, odr-7, and AWC(-) worms for (D) isoamyl alcohol (IAA), (E) acetoin (ACE),

(F) isovalerate (ISV), (G) isobutanol (ISB), (H) phenylethyl alcohol (PEA), (I) butyl acetate (BA), and (J) methyl isovalerate (MIV)

(K) Summary schematic representing the role of AWA and AWC odor sensory neurons in ‘odor-only’ diet preference for leucine-enriched diets and chemotaxis assays.

Significant differences are indicated as * P ≤ 0.05, ** P ≤ 0.01, *** P ≤ 0.001, and **** P ≤ 0.0001 determined by one-way ANOVA followed by post hoc Dunnett’s multiple comparison test. Error bars indicate SEM (n=15).

To identify the constituents of foraging signal, we sampled the headspace of each CeMbio bacteria. Each bacterium produced a unique bouquet of odors composed of chemically diverse molecules, including alcohols, aldehydes, ketones, esters, and carboxylic acids (Table S1). To identify foraging cues amongst odors in the headspace of preferred bacteria, we examined the response of WT worms to ten different odors in chemotaxis assays (schematic in Figure S5A). We found that WT worms were attracted to IAA, acetoin (ACE), isovalerate (ISV), isobutanol (ISB), and phenylethyl alcohol (PEA) in a dose-dependent manner (Figures S5B-S5F). Worms showed attraction to butyl acetate (BA) and methyl isovalerate (MIV) only at specific concentrations but did not display a clear dose-response (Figures S5G and S5H). Worms showed no response to 2,3-butanediol (BDL), isoamyl acetate (ISA), and indole (IND) (Figures S5I-S5K). This is the first report of acetoin and methyl isovalerate as attractants for C. elegans. These results show that the headspace of preferred bacteria contains multiple attractive odors for C. elegans.

To identify neurons responsible for sensing attractive odors produced by preferred bacteria, we used the odr-7 and AWC(-) worms in chemotaxis assays for each of the seven attractive odors. We found that AWC(-) worms had severely diminished responses to IAA, ISV, ISB, PEA, BA, and MIV (Figures 3D and 3F-3J). Acetoin, on the other hand, was sensed by AWA neurons as odr-7 showed reduced response to acetoin (Figure 3E). Taken together, our findings revealed that worms sense an extensive repertoire of chemically diverse odors produced by preferred bacteria as attractants, predominantly using AWC neurons (summary in Figure 3K).

Isoamyl alcohol drives foraging behavior in C. elegans

To identify the most relevant foraging signal for C. elegans, we took advantage of the well-established odor adaptation assay using each of the attractive odors in its diets (schematic in Figure 4A). Prolonged exposure of C. elegans to an odorant is known to result in the loss or reduction of the ability of worms to sense the same odorant in a subsequent exposure 53. We performed a modified adaptation assay where worms were exposed to a bouquet of odors from a bacterial lawn followed by testing their chemotaxis response to individual attractive odors. Worms adapted to odors from CEent1 lawns had dramatically reduced responses to IAA, isobutanol, and phenylethyl alcohol compared to naïve worms (Figures 4B, 4E, and 4F). Worms adapted to odors from JUb66 and BIGb0170 lawns also had dramatically reduced response to IAA (Figure 4B). While adaptation to bacterial odors did not influence worms’ response to acetoin, isovalerate, butyl acetate and methyl isovalerate (Figure 4C, 4D, 4G and 4H). These findings suggested that the odor bouquets of preferred bacteria have levels of IAA high enough to modulate chemoperception in C. elegans.

Isoamyl alcohol regulates the diet preference of worms.

(A) Schematic representation of the adaptation regimen followed by chemotaxis assays.

Chemotaxis index (CI) for naïve worms and worms adapted with CEent1, JUb66, or BIGb0170 odors to (B) isoamyl alcohol (IAA), (C) acetoin (ACE), (D) isovalerate (ISV), (E) isobutanol (ISB) (F) phenylethyl alcohol (PEA), (G) butyl acetate (BA), and (H) methyl isovalerate (MIV). Dark gray bars indicate naïve worms and light gray bars indicate worms adapted to bacterial odors.

Significant differences are indicated as * P ≤ 0.05, ** P ≤ 0.01, *** P ≤ 0.001, and **** P ≤ 0.0001 determined by one-way ANOVA followed by post hoc Dunnett’s multiple comparison test. Error bars indicate SEM (n=15).

If IAA is the most relevant foraging signal, we predicted that prior adaptation to IAA will diminish the diet preference of C. elegans. To test this, we adapted worms to IAA odor and checked their diet preference (schematic in Figure S6A). IAA-adapted worms showed a reduced chemotaxis response to IAA, indicating that the adaptation regimen is robust (Figure S6B). We found that IAA-adapted worms displayed a reduced preference for CEent1, JUb66, or BIGb0170 over E. coli OP50 compared to naïve worms (Figures S6C-S6E). These results indicated that IAA is the predominant olfactory cue that determines the diet preference of C. elegans.

IAA is an ecologically relevant odor for C. elegans

If an odor is relevant as a foraging cue in nature, the ability to sense it must be under positive selection in wild C. elegans populations. We anticipated that most wild isolates of C. elegans will have a robust response to IAA but not to other attractants reported in this study. To test this hypothesis, we studied the response to IAA in nine wild isolates of C. elegans collected from geographically distinct regions (Figure 5A and Table S1) 54,55. Additionally, we tested their chemotaxis response to diacetyl, a known food cue, and two other attractants (PEA and ACE) identified in this study. We found that all the strains displayed robust response to IAA comparable to WT worms (Figure 5B). The chemotaxis response of these strains to diacetyl was also quite robust (Figure 5C). However, the chemotaxis response to PEA and acetoin were variable within and across wild isolates suggesting they are not crucial for foraging (Figures 5D and 5E). The robust response of wild isolates to IAA supports the notion that the ability of C. elegans to sense IAA has been selected during evolution. These findings suggest that IAA is an ecologically relevant foraging cue for C. elegans.

Robust chemoperception of isoamyl alcohol in wild isolates of C. elegans.

(A) World map representing the distinct geographical locations (source) of wild isolates of C. elegans used in this study.

Chemotaxis index (CI) of WT and wild isolates of worms for (B) IAA, (C) diacetyl (DA), (D) phenylethyl alcohol (PEA), and (E) acetoin (ACE).

Significant differences are indicated as * P ≤ 0.05, ** P ≤ 0.01, *** P ≤ 0.001, and **** P ≤ 0.0001 determined by one-way ANOVA followed by post hoc Dunnett’s multiple comparison test. Error bars indicate SEM (n=15).

SRD-12 senses isoamyl alcohol and facilitates foraging in C. elegans

C. elegans genome encodes for ∼1300 putative GPCRs, many of which are predicted to sense soluble and olfactory cues 56. Since the preference for IAA and leucine-enriched diets is mediated by AWC odor sensory neurons (Figures 3A-3D), we hypothesized that the receptor for IAA is a GPCR expressed in AWC neurons. Indeed, calcium response to IAA has been reported in AWC neurons 57,58. Using single-cell RNAseq data for C. elegans neurons, available at CeNGEN 44, we listed 18 putative GPCRs expressed in AWCon and AWCoff for further analysis (Figure 6A). We tested response to IAA in 15 strains harboring mutants in one or more of these GPCRs (see Table S1) 59. One of these strains, CHS1169, had a significantly reduced response to IAA compared to WT worms (Figure 6A). Of the six GPCRs edited in the strain, srd-12 had the highest expression in AWCoff odor sensory neuron 44. Using CRISPR/Cas9-based approach, we generated two srd-12 deletion lines, VSL2401, and VSL2402, carrying mutation in srd-12 gene (see methods and Table S1). We recorded the chemotaxis response of WT and mutant worms to IAA (schematic in Figure 6B) and found they were indeed defective in their response to IAA (Figures 6C-6F). By analyzing the locomotion of WT, VSL2401, and VSL2402 worms, in the chemotaxis arena, we determined quantitative differences in their response to IAA (Figure S7). We observed that VSL2401 and VSL2402 strains had speeds comparable to WT worms and traveled distances similar to WT worms suggesting that srd-12 mutants do not have locomotory defects (Figures S7A and S7B). We found that both srd-12 mutant strains had significantly reduced chemotaxis response to IAA compared to WT worms (Figure 6F). However, srd-12 mutants had a comparable response to WT worms for acetoin and phenylethyl alcohol (Figures 6G and 6H). As expected, srd-12 mutant strains had diminished preference for CEent1 over E. coli OP50 in the diet preference assay consistent with the idea that IAA is the most relevant foraging signal (Figure 6I). Taken together, our findings show that C. elegans senses foraging signal, IAA, using newly identified GPCR, SRD-12.

SRD-12 GPCR mediates isoamyl alcohol sensing and diet preference in C. elegans.

(A) List of GPCRs that are highly expressed in AWC neurons. Chemotaxis index (CI) of WT and GPCR edited strains to 1:1000 IAA (n≥3).

(B) Schematic for chemotaxis assay plate used for movement track analysis.

Movement tracks of 5 animals in a chemotaxis arena with 1:1000 IAA for 15 minutes, for (C) WT, (D) VSL2401, and (E) VSL2402. Also, refer to the video S2. Individual color represents the track for a single worm.

Chemotaxis index (CI) of WT, VSL2401, and VSL2402 worms for (F) 1:1000 IAA, (G) 500 mM acetoin (ACE), and (H) absolute phenylethyl alcohol (PEA).

(I) Preference index (PI) of WT, VSL2401, and VSL2402 worms for a preferred diet, CEent1, over E. coli OP50. symbol indicates ‘odor-only’ preference assays.

Significant differences are indicated as * P ≤ 0.05, ** P ≤ 0.01, and **** P ≤ 0.0001 determined by one-way ANOVA followed by post hoc Dunnett’s multiple comparison test. Error bars indicate SEM (n≥15).

Based on our study, we propose a model describing an odor-dependent mechanism for sensing EAA-enriched bacteria in C. elegans. We show that IAA is an olfactory signal for leucine-enriched diet. AWC olfactory neurons of C. elegans allow worms to choose for IAA producing bacteria. Bacteria preferred by worms have the metabolic capacity to produce IAA from leucine via the Ehrlich degradation pathway. SRD-12, a GPCR in AWC neurons, mediates foraging for bacteria in C. elegans by facilitating chemoperception of IAA (Figure 7).

Model depicting odor-based foraging strategy used by C. elegans.

C. elegans prefers leucine-enriched diets in an odor-dependent manner. Preferred bacteria catabolize leucine, an EAA, to produce IAA via the Ehrlich degradation pathway. SRD-12 GPCR expressed in AWC odor sensory neurons mediates the foraging behavior of C. elegans by sensing IAA.

Discussion

Our study provides evidence for an odor-dependent mechanism for foraging in C. elegans. Using CeMbio, a model microbiome for C. elegans, we show that worms prefer leucine supplemented bacteria, namely Enterobacter hormaechei, Lelliottia amnigena, and Sphingobacterium multivorum, in an odor-dependent manner. Leucine-enriched diets produce significantly higher levels of IAA odor, making up to 90% of their headspace. These bacteria produce IAA from leucine using the Ehrlich degradation pathway. Preference for leucine-enriched bacteria solely depends on the AWC odor sensory neurons of C. elegans. Although the preferred bacteria produce several AWC-sensed odors, IAA exclusively influences the diet preference of worms. Wild isolates of C. elegans, representing nine geographically distinct locations on Earth, respond robustly to IAA underscoring its relevance as a foraging signal in nature. C. elegans utilizes SRD-12, a GPCR primarily expressed in AWC neurons, to sense IAA and forage for the preferred bacteria. Thus, IAA-SRD-12 represents a ligand-receptor module used by C. elegans to forage for bacteria in its natural environment.

Is olfaction a major contributor to foraging behavior in animals in general? Artificial flowers, containing nectars with high amino acid content, are better at attracting butterflies 60. Honeybees prefer to consume artificial nectar rich in proline 61. Hummingbirds, moths, and ants are attracted to flowers emitting benzyl acetone, an odor synthesized from an EAA, phenylalanine 62–65. For odors to serve as an attractant, the said odor must serve as a proxy or code for a needed primary metabolite, abundant in carbon, nitrogen, phosphorus, or micronutrients. In such a scenario, it is easy to envisage that each attractive cue might be a code for a specific essential nutrient.

Is isoamyl alcohol a signal for essential nutrient for C. elegans and other foraging animals. IAA was one of 121 odors reported to be sensed by C. elegans several decades ago 47, but its ecological relevance was unclear to date. Our study suggests that IAA is the primary cue that drives foraging in favor of leucine-enriched diets in C. elegans. Does IAA reflect the metabolic state of bacteria producing it? IAA is indeed derived from a primary metabolite leucine, an essential branched chain amino acid for animals 66. When bacteria were provided with additional leucine, the absolute abundance of IAA increases significantly in three different bacteria examined (Figures 2C and S3). The arrival of feeding insects to ripening fruits such as bananas coincides with for production of IAA and isoamyl acetate in the fruit via Ehrlich degradation pathway 67. The presence of Ehrlich degradation pathway in specific microbes and many insect attracting plants underscores the importance of odors in mediating interkingdom interactions. IAA, in combination with other odors, has been reported to attract Drosophila melanogaster and help in establishing a fly-fruit-yeast relationship 68–74. Why do animals need BCAA? Not only is leucine a major component of C. elegans proteome, intracellular accumulation of BCAA including leucine prolongs the lifespan of C. elegans 75. The lack of direct leucine perception (Figure S1H in this study) but presence of IAA perception ability, via SRD-12, suggests that worms rely on IAA sensing to identify bacteria with leucine. Additional natural microbiotas of C. elegans (not part of CeMbio) which produce IAA have also been reported to attract worms 76,77. Worms also prefer odors derived from other BCAA such as isobutanol derived from valine via Ehrlich degradation (Figure 4E) 78–80. Phenylethyl alcohol (PEA in this study) can be produced by microbial degradation of phenyl alanine, another essential amino acid that animals require for their proteins as well as for the synthesis of neurotransmitter dopamine. All these reports support the idea that animals use olfactory cues to identify diets rich in essential amino acids.

Animals use GPCRs to sense soluble and olfactory cues present in their environment. Rats utilize taste receptors T1R1/T1R3 for sensing alanine 81. C. elegans can also sense lysine and histidine 82 but not leucine (this study). Several studies have implicated olfactory GPCRs in regulating animal behavior. In termites, olfactory co-receptor OCRO influences foraging by enabling pheromone sensing 83. Or83b, an odorant receptor in D. melanogaster regulates lifespan 84. Six odor sensory neurons of C. elegans express hundreds of GPCRs, most of them remain orphan 44.ODR-10 expressed in AWA odor sensory neurons regulates foraging behavior in C. elegans for lactic acid bacteria which produce diacetyl odor 85. AWC odor sensory neurons express as many as 171 GPCRs, of which 18 GPCRs display strong expression. These receptors likely mediate the foraging behavior of worms. One of these, STR-2, senses 2-heptanone odor produced by Bacillus nematocidal 86. STR-2 also regulates lipid accumulation and lifespan in a temperature-dependent manner 87. The ligand(s) for the remaining 17 receptors were unknown. The discovery of one of these, SRD-12, as receptor for sensing IAA (this study) suggests that the ligand for other receptors may also be found in the headspace of the C. elegans microbiome. Our study provides a framework for the identification of ligands for orphan GPCR, by searching an ecologically relevant repertoire, the microbiome of an animal.

Methods

Experimental model and subject details Strains and growth media

C. elegans strains used in this study are listed in see Table S1. All strains were maintained as hermaphrodites at 20°C on nematode growth media (NGM) plates seeded with E. coli OP50, as originally described 88. All the CeMbio strains used in this study (see Table S1) were grown on Difco Luria-Bertani (LB) media and maintained at 25°C, as described by Dirksen et al. 48. A list of all the reagents used in this study is provided in Table S1.

CRISPR approach for gene editing

Homology-directed integration of the ssDNA oligo was done in the srd-12 gene using a CRISPR/Cas9-based approach as described previously 89. ssDNA oligo was designed to contain two stop codons along with a restriction site between two 35 bp long homology arms flanking the PAM sites of the targeted gene. ssDNA oligo was incubated with optimized RNP complexes of Cas9, crRNA, and tracrRNA. The mix was injected in the gonad of C. elegans along with rol-6 [pRF4::rol-6(su1006)] marker plasmid. The F1 progenies were segregated and genotyped using PCR, followed by restriction digestion for the introduction of mutation. srd-12 CRISPR mutants were back-crossed three times with N2 Bristol WT strain to generate VSL2401 and VSL2402.

Diet preference assay

For diet preference assays, individual CeMbio bacterium (test bacterium) and E. coli (control bacterium) were grown in LB broth overnight at 25°C and adjusted to an OD600 of 1. Standard diet choice assays were conducted on 90 mm NGM plates, wherein 25 µl of cultures of the test and control bacteria were spotted 1 cm away from the periphery of the plate on diametrically opposite points of the dish. These plates were incubated at 25°C for 12 hours. Gravid adult worms were washed thrice with S-basal buffer (100 mM NaCl, 5.75 mM K2HPO4, 44 mM KH2PO4). 50-70 worms were placed in the center of the assay plate and incubated at 25°C. The preference index (PI) was calculated after 3 hours using the following equation:

Preference Index (PI) = (Worms on the test lawn – Worms on the control lawn) / Total worms

‘Odor-only’ diet preference assays were performed in tripartitioned plates, wherein one compartment was filled with buffered agar or BA (1 mM CaCl2, 1 mM MgSO4, 5 µg/ml cholesterol, 25 mM KPO4 buffer, and 2% agar) and other two compartments were filled with NGM. All the assay plates were air-dried for 70-90 minutes. For EAA supplementation, one of the NGM-filled compartments was supplemented with individual EAA at 5 mM concentration (+ EAA), while the other was left unsupplemented (-EAA). All bacterial strains were grown in LB broth overnight at 25°C and adjusted to an OD600 of 1. A volume of 25 µl of the test bacterial culture was spotted on the two compartments with NGM. The plates were incubated at 25°C for 12 hours. At the time of assay, 1 µl of sodium azide was spotted near the edges of the BA-containing compartment. Gravid adult worms were washed thrice with S-basal buffer, and 50-70 worms were placed at the center of the BA-containing compartment and incubated at 25°C (schematic in Figure 1A). The preference index (PI) was calculated after 3 hours using the following equation:

Preference index (PI) = [(Worms near diet+EAA) – (Worms near diet-EAA)] / Total worms

To understand worms’ preference for individual CeMbio bacterium (test bacterium) over E. coli OP50 (control bacterium), we performed an ‘odor-only’ diet preference assay as described above without any supplementations. The test bacterium culture was spotted in one of the NGM-filled compartments, while E. coli OP50 was spotted in the other compartment (schematic in Figure S2A). The remaining steps of the assay were performed in a manner similar to the one mentioned above. The PI was calculated using the following equation:

Preference index (PI) = (Worms near CeMbio – Worms near OP50) / Total worms

All diet choice experiments were conducted as standard preference assays unless otherwise specified.

Identification of volatiles produced by microbiota of C. elegans

The volatile profiles of CeMbio strains and E. coli OP50 were determined using gas chromatography-mass spectrometry (GC-MS/MS) 90. Briefly, the individual bacterium was inoculated in 3 ml of LB broth and grown overnight at 25°C. NGM dishes (60 mm) were seeded with seven spots of 50 µl bacterial cultures adjusted to an OD600 of 1. Seeded plates were then incubated at 25°C for 23 hours. Two plates of the same conditions were sealed together using parafilm and incubated at 25°C for 1 hour to build odor concentration in the headspace for sampling. The volatiles were collected using a solid-phase microextraction fiber (SPME DVB: Divinylbenzene – Carboxen-WR – polydimethyl siloxane, 80 µm: Agilent Technologies, Part no. 5191-5874). Odors were subjected to thermal desorption and were identified using gas chromatography-mass spectrometry in an 8890C gas chromatography machine coupled with a 7000D GC/TQ, which used a capillary column HP-5MS ultra inert (30 m x 0.25 mm and 0.25 m, Agilent 19091S-433UI: 0245625H) with Helium as the carrier gas at a constant flow rate of 1.5 ml/min. The injection of the sample was done at 40°C. The GC program included a 40°C hold for 1 minute followed by a temperature ramp to 170°C at a rate of 5°C/min, then by a ramp to 270°C at a rate of 100°C/min, and finally, hold for 2 minutes. The inlet, M.S. source, and M.S. quadrupole temperatures were maintained at 225°C, 230°C, and 150°C, respectively. Odors were then identified using NIST (National Institute of Standards and Technology) 2019 V2.3 Mass Spectral Library and Agilent MassHunter Workstation version 10.0. The m/z peaks detected in the odor profile of NGM plates seeded with LB broth were subtracted from the odor profile of each tested bacterium. For leucine supplementation, preferred diets (CEent1, JUb66, and BIGb0170) were seeded on NGM supplemented with 5 mM leucine (+ LEU) and without supplementation (-LEU). GC-MS/MS was performed as described above, and a heat map was generated by analyzing the area under the curve of individual volatile peaks (Figure 2C).

For quantification of IAA, a standard curve was prepared by measuring the area under the curve for five concentrations (0.18 µM, 0.36 µM, 2.25 µM, 4.50 µM, and 8.99 µM) of IAA (Figure S3E). The preferred diets were seeded with two spots of 50 µl culture on NGM dishes supplemented with (+ LEU) and without (-LEU) leucine. The amount of IAA produced by each bacterium was determined by comparing the area under the curve in the GC-MS/MS plot of each bacterium against the standard curve for IAA.

Bacterial RNA extraction and qRT-PCR relative expression analysis

For RNA isolation, bacterial culture was grown in 20 ml of NGM broth with 5 mM leucine (+ LEU) and without leucine (-LEU) for 24 hours at 25°C. Bacterial cells were harvested by centrifugation at 5000 rpm for 20 min. The bacterial pellet was treated with RNAprotect Bacteria Reagent (Qiagen, Cat. No. 76506). Total RNA was extracted from the samples using the RNeasy Mini kit (Qiagen, Cat. No. 74104), and DNase I (NEB Cat. No. M0303S) treatment was done to remove genomic DNA. cDNA was synthesized using the iScript cDNA synthesis kit (Bio-Rad, Cat. No. 170-8891) and used for qRT-PCR to analyze the relative gene expression of ilvE using SYBR Green detection mix (Bio-Rad Cat. No. 1725124) on StepOnePlus (Applied Biosystems) machine. rboB was used as a housekeeping gene. The comparative ΔΔCt method was used to determine the fold change of the ilvE target gene.

Chemotaxis assay

Chemotaxis assays were performed as described previously 90. Briefly, BA plates (90 mm) air-dried for 90 minutes were used for the assays. Sodium azide (2 µl) was spotted on the two diametrically opposite ends of the plate, followed by 2 µl of the test chemical to one side and solvent to the opposite side (schematic in Figure S4A). Test chemicals were diluted using suitable solvents (Table S1). Gravid adult worms were washed thrice with S-basal buffer. 50 to 80 gravid adult worms were placed in the center of the plate and incubated at 25°C for 3 hours. The chemotaxis index (CI) was calculated using the following equation:

Chemotaxis index (CI) = (Worms towards test – Worms towards control) / Total worms

Odor adaptation

Gravid adult worms were washed thrice with S-basal buffer and transferred to 60 mm NGM plates. For odor adaption, worms were exposed to bacterial odors by sealing an NGM plate seeded with bacteria (7 spots of 50 µL bacterial culture) with the plate containing washed worms. These plates were incubated at 25°C for 90 minutes to desensitize the worms to the odor. After exposure, worms were washed once with S basal buffer and used to perform chemotaxis assays (schematic in Figure 4A). For the modified odor adaptation assays, the NGM plate containing worms was exposed to IAA (1:10 dilution) by sealing it with another NGM agar plate containing 4 spots of 3 µl IAA or without IAA (schematic in Figure S6A). The adapted worms were then used for diet preference or chemotaxis assays as required.

Worm movement track analysis

We recorded the motion of five worms on a chemotaxis assay plate using a simple imaging setup. We used 1800 frames (1 fps) to track the behavior of worms. Using MATLAB, we extracted the set of binary images from the recorded frames and processed them using the Trackmate plugin of Fiji software to track worms’ motion. We used the Advanced Kalman algorithm to create tracks and interpolate for missing frames. The trajectory information was exported in the motilitylab spreadsheet format. This data was analyzed using MATLAB to measure the average speed of worms and the total distance traveled by worms 91–94.

Statistical analyses

All statistical analyses were done using GraphPad Prism version 8. Statistical analyses were performed either by a two-tailed unpaired t-test between two groups or one-way ANOVA, followed by post hoc Dunnett’s multiple comparison test across multiple groups. The significant differences were denoted according to P-values: *P ≤ 0.1; **P ≤ 0.01; ***P ≤ 0.001; and ****P ≤ 0.0001. Data are presented as means ± SEM. All experiments were performed with 15 biological replicates done over three days.

eLife Assessment

This is an important study, supported by solid data, that suggests a model for diet selection in C. elegans. The significance is that while C. elegans has long been known to be attracted to bacterial volatiles, what specific bacterial volatiles may signify to C. elegans is largely unknown. This study also provides evidence for a possible odorant/GPCR pairing.

Reviewer #1 (Public review):

Summary:

Siddiqui et al., investigate the question of how bacterial metabolism contributes to the attraction of C. elegans to specific bacteria. They show that C. elegans prefers three bacterial species when cultured in a leucine-enriched environment. These bacterial species release more isoamyl alcohol, a known C. elegans attractant, when cultured with leucine supplement than without leucine supplement. The study shows correlative evidence that isoamyl alcohol is produced from leucine by the Ehrlich pathway. In addition, they show that SRD-12 is likely a receptor for isoamyl alcohol because a null mutant of this receptor exhibits lower chemotaxis to isoamyl alcohol and lower preference for leucine-enriched bacteria.

Strengths:

(1) This study takes a creative approach to examine the question of what specific volatile chemicals released by bacteria may signify to C. elegans by examining both bacterial metabolism and C. elegans preference behavior. Although C. elegans has long been known to be attracted to bacterial metabolites, this study may be one of the first to examine the role of a specific bacterial metabolic pathway in mediating attraction.

(2) A strength of the paper is the identification of SRD-12 as a likely receptor for isoamyl alcohol. The ligands for very few olfactory receptors have been identified in C. elegans and so this is a significant addition to the field. The srd-12 null mutant strain will likely be a useful reagent for many labs examining olfactory and foraging behaviors.

Weaknesses:

(1) The authors write that the leucine metabolism via the Ehrlich pathway is required for the production of isoamyl alcohol by three bacteria (CEent1, JUb66, BIGb0170), but their evidence for this is correlation and not causation. They write that the gene ilvE is a bacterial homolog of the first gene in the yeast Ehrlich pathway (it would be good to include a citation for this) and that the gene is present in these three bacterial strains. In addition, they show that this gene, ilvE, is upregulated in CEent1 bacteria upon exposure to leucine. To show causation, they need to knockout ilvE from one of these strains, show that the bacteria does not have increased isoamyl alcohol production when cultured on leucine, and that the bacteria is no longer attractive to C. elegans.

(2) The authors examine three bacterial strains that C. elegans showed increased preference when grown with leucine supplementation vs. without leucine supplementation. However, there also appears to be a strong preference for another strain, JUb0393, when grown on plus leucine (Figure 1B). It would be good to include statistics and criteria for selecting the three strains.

3. Although the behavioral evidence that srd-12 gene encodes a receptor for isoamyl alcohol is compelling, it does not meet the standard for showing that it is an olfactory receptor in C. elegans. To show it is indeed a likely receptor one or more of the following should be done:

(a) Calcium imaging of AWC neurons in response to isoamyl alcohol in the receptor mutant with the expectation that the response would be reduced or abolished in the mutant compared to wildtype.

(b)"A receptor swap" experiment where the SRD-12 receptor is expressed in AWB repulsive neuron in SRD-12 receptor mutant background with the expectation that with receptor swap C. elegans will now be repulsed from isoamyl alcohol in chemotaxis assays (experiment from Sengupta et al., 1996 odr-10 paper).

(4) The authors conclude that C. elegans cannot detect leucine in chemotaxis assays. It is important to add the method for how leucine chemotaxis assay was done in order to interpret these results. Because leucine is not volatile if leucine is put on the plates immediately before the worms are added (as in a traditional odor chemotaxis assay), there is no leucine gradient for the worm to detect. It would be good to put leucine on the plate several hours before worms are introduced so worms have the possibility to be able to detect the gradient of leucine (for example, see Wakabayashi et al., 2009).

(5) The bacterial preference assay entitled "odor-only assay" is a misleading name. In the assay, C. elegans is exposed to both volatile chemicals (odors) and non-volatile chemicals because the bacteria are grown on the assay plate for 12 hours before the worms are introduced to the assay plate. In that time, the bacteria is likely releasing non-volatile metabolites into the plate which may affect the worm's preference. A true odor-only assay would have the bacteria on the lid and the worms on the plate.

(6) The findings of the study should be discussed more in the context of prior literature. For example, AWC neurons have been previously shown to be involved in bacterial preference (Harris et al., 2014; Worthy et al., 2018). In addition, CeMbio bacterial strains (the strains examined in this study) have been previously shown to release isoamyl alcohol (Chai et al. 2024).

Reviewer #2 (Public review):

Summary:

Siddiqui et al. show that C. elegans prefers certain bacterial strains that have been supplemented with the essential amino acid (EEA) leucine. They convincingly show that some leucine enriched bacteria stimulate the production of isoamyl alcohol (IAA). IAA is an attractive odorant that is sensed by the AWC. The authors an identify a receptor, SRD-12, that is expressed in the AWC chemosensory neurons and is required for chemotaxis to IAA. The authors propose that IAA is a predominant olfactory cue that determines diet preference in C. elegans. Since leucine is an EAA, the authors propose that worm IAA sensing allows the animal provides a proxy mechanism to identify EAA rich diets.

Strengths:

The authors propose IAA as a predominant olfactory cue that determines diet preference in C. elegans providing molecular mechanism underlying diet selection. They show that wild isolates of C. elegans have a strong chemotactic response to IAA indicating that IAA is an ecologically relevant odor for the worm. The paper is well written, and the presented data are convincing and well organized. This is an interesting paper that connects chemotactic response with bacterially produced odors and thus provides an understanding of how animals adapt their foraging behavior through the perception of molecules that may indicate the nutritional value.

Weaknesses:

Major:

While I do like the way the authors frame C. elegans IAA sensing as mechanisms to identify leucine (EAA) rich diets it is not fully clear whether bacterial IAA production is a proxy for bacterial leucine levels.

(1) Can the authors measure leucine (or other EAA) content of the different CeMbio strains? This would substantiate the premise in the way they frame this in the introduction. While the authors convincingly show that leucine supplementation induces IAA production in some strains, it is not clear if there are lower leucine levels in the different in non-preferred strains.

(2) It is not clear whether the non-preferred bacteria in Figure 1A and 1B have the ability to produce IAA. To substantiate the claim that C. elegans prefers CEent1, JUb66, and BIGb0170 due to their ability to generate IAA from leucine, it would measure IAA levels in non-preferred bacteria (+ and - leucine supplementation). If the authors have these data it would be good to include this.

(3) The authors would strengthen their claim if they could show that deletion or silencing ilvE enzyme reduces IAA levels and eliminates the increased preference upon leucine supplementation.

(4) While the three preferred bacteria possess the ilvE gene, it is not clear whether this enzyme is present in the other non-preferred bacterial strains. As far as I know, the CeMbio strains have been sequenced so it should be easy to determine if the non-preferred bacteria possess the capacity to make IAA. Does the expression of ilvE in e.g. E. coli increase its preference index or are the other genes in the biosynthesis pathway missing?

(5) It is strongly implied that leucine-rich diets are beneficial to the worm. Do the authors have data to show the effect on leucine supplementation on C. elegans healthspan, life-span or broodsize?

Other comments:

Page 6. Figure 2c. While the authors' conclusions are correct based on AWC expts. it would be good at this stage to include the possibility that odors that enriched in the absence of leucine may be aversive.

Page 6. IAA increases 1.2-4 folds upon leucine supplementation. If the authors perform a chemotaxis assay with just IAA with 1-2-4 fold differences do you get the shift in preference index as seen with the bacteria? i.e. is the difference in IAA concentration sufficient to explain the shift in bacterial PI upon leucine supplementation? Other attractants such as Acetoin and isobutanol go up in -Leu conditions.

Page 14-15. The authors identify a putative IAA receptor based on expression studies. I compliment the authors for isolating two CRISPR deletion alleles. They show that the srd-12 mutants have obvious defects in IAA chemotaxis. Very few ligand-odorant receptors combinations have been identified so this is an important discovery. CenGen data indicate that srd-12 is expressed in a limited set of neurons. Did the authors generate a reporter to show the expression of srd-12? This is a simple experiment that would add to the characterization of the SRD-12 receptor. Rescue experiments would be nice even though the authors have independent alleles. To truly claim that SRD-12 is the ligand for IAA and activates the AWC neurons would require GCamp experiments in the AWC neuron or heterologous expression system. I understand that GCamp imaging might not be part of the regular arsenal of the lab but it would be a great addition (even in collaboration with one of the many labs that do this regularly). Comparing AWC activity using GCaMP in response IAA-producing bacteria with high leucine levels in both wild-type and SRD-12-deficient backgrounds, would further support their narrative. I leave that to the authors.

Minor:

Page 4 "These results suggested that worms can forage for diets enriched in specific EAA, leucine...." More precise at this stage would be to state " These results indicated that worms can forage for diets supplemented with specific EAA...".

Page 5."these findings suggested that worms not only rely on odors to choose between two bacteria but also to find leucine enriched bacteria" This statement is not clear to me and doesn't follow the data in Fig. S2. Preferred diets in odorant assays are the IAA producing strains.

Page 5. Figure S2A provides nice and useful data that can be part of the main Figure 1.

Reviewer #3 (Public review):

Summary:

The authors first tested whether EAA supplementation increases olfactory preference for bacterial food for a variety of bacterial strains. Of the EAAs, they found only leucine supplementation increased olfactory preference (within a bacterial strain), and only for 3 of the bacterial strains tested. Leucine itself was not found to be intrinsically attractive.

They determined that leucine supplementation increases isoamyl alcohol (IAA) production in the 3 preferred bacterial strains. They identify the biochemical pathway that catabolizes leucine to IAA, showing that a required enzyme for this pathway is upregulated upon supplementation.

Consistent with earlier studies, they find that AWC olfactory neuron is primarily responsible for increased preference for IAA-producing bacteria.

Testing volatile compounds produced by bacteria and identified by GC/MS, and identified several as attractive, most of them require AWC for the full effect. Adaptation assays were used to show that odorant levels produced by bacterial lawns were sufficient to induce olfactory adaptation, and adaptation to IAA reduced chemotaxis to leucine-supplemented lawns. They then showed that IAA attractiveness is conserved across wild strains, while other compounds are more variable, suggesting IAA is a principal foraging cue.

Finally, using the CeNGEN database, they developed a list of candidate IAA receptors. Using behavioral tests, they show that mutation of srd-12 greatly impairs IAA chemotaxis without affecting locomotion or attraction to another AWC-sensed odor, PEA.

Comments

This study will be of great interest in the field of C. elegans behavior, chemical senses and chemical ecology, and understanding of the sensory biology of foraging.

Strengths:

The identification of a receptor for IAA is an excellent finding. The combination of microbial metabolic chemistry and the use of natural bacteria and nematode strains makes an extremely compelling case for the ecological and adaptive relevance of the findings.

Weaknesses:

AWC receives synaptic input from other chemosensory neurons, and thus could potentially mediate navigation behaviors to compounds detected in whole or in part by those neurons. Language concluding detection by AWC should be moderated (e.g. p9 "worms sense an extensive repertoire...predominantly using AWC") unless it has been demonstrated.

srd-12 is not exclusively expressed in AWC. Normally, cell-specific rescue or knockdown would be used to demonstrate function in a specific cell. The authors should provide such a demonstration or explain why they are confident srd-12 acts in AWC.

A comparison of AWC's physiological responses between WT and srd-12, preferably in an unc-13 background, would be nice. Even further, the expression of srd-12 in a different neuron type and showing that it confers responsiveness to IAA (in this case, inhibition) would be very convincing.

Author response:

The following is the authors’ response to the original reviews.

eLife Assessment:

This is an important study, supported by solid to convincing data, that suggests a model for diet selection in C. elegans. The significance is that while C. elegans has long been known to be attracted to bacterial volatiles, what specific bacterial volatiles may signify to C. elegans is largely unknown. This study also provides evidence for a possible odorant/GPCR pairing.

Public Reviews:

Reviewer #1 (Public review):

Summary:

Siddiqui et al., investigate the question of how bacterial metabolism contributes to the attraction of C. elegans to specific bacteria. They show that C. elegans prefers three bacterial species when cultured in a leucine-enriched environment. These bacterial species release more isoamyl alcohol, a known C. elegans attractant, when cultured with leucine supplement than without leucine supplement. The study shows correlative evidence that isoamyl alcohol is produced from leucine by the Ehrlich pathway. In addition, they show that SRD-12 (SNIF-1) is likely a receptor for isoamyl alcohol because a null mutant of this receptor exhibits lower chemotaxis to isoamyl alcohol and lower preference for leucine-enriched bacteria.

Strengths:

(1) This study takes a creative approach to examine the question of what specific volatile chemicals released by bacteria may signify to C. elegans by examining both bacterial metabolism and C. elegans preference behavior. Although C. elegans has long been known to be attracted to bacterial metabolites, this study may be one of the first to examine the role of a specific bacterial metabolic pathway in mediating attraction.

(2) A strength of the paper is the identification of SRD-12 (SNIF-1) as a likely receptor for isoamyl alcohol. The ligands for very few olfactory receptors have been identified in C. elegans and so this is a significant addition to the field. The srd-12 (snif-1) null mutant strain will likely be a useful reagent for many labs examining olfactory and foraging behaviors.

Weaknesses:

(1) The authors write that the leucine metabolism via the Ehrlich pathway is required for the production of isoamyl alcohol by three bacteria (CEent1, JUb66, BIGb0170), but their evidence for this is correlation and not causation. They write that the gene ilvE is a bacterial homolog of the first gene in the yeast Ehrlich pathway (it would be good to include a citation for this) and that the gene is present in these three bacterial strains. In addition, they show that this gene, ilvE, is upregulated in CEent1 bacteria upon exposure to leucine. To show causation, they need to knockout ilvE from one of these strains, show that the bacteria does not have increased isoamyl alcohol production when cultured on leucine, and that the bacteria is no longer attractive to C. elegans.

Thank you for the comment. We have added the appropriate citation [1,2]. We agree that worms’ diet preference for the preferred strains upon ilvE knockout will further strengthen the claim for IAA being used as a proxy for leucine-enriched diet. Currently, protocols and tools for genetic manipulations for CeMbio strains are not available, making this experiment not feasible at this time.

(2) The authors examine three bacterial strains that C. elegans showed increased preference when grown with leucine supplementation vs. without leucine supplementation. However, there also appears to be a strong preference for another strain, JUb0393, when grown on plus leucine (Figure 1B). It would be good to include statistics and criteria for selecting the three strains.

Thanks for your comment. We agree that for Pantoea nemavictus, JUb393, worms seem to prefer the leucine supplemented (+ LEU) bacteria over unsupplemented (-LEU). However, when given a choice between the individual CeMbio bacteria and E. coli OP50, worms showed preference for only CEent1, JUb66, and BIGb0170 (Figure 1F). Consequently, CEent1, JUb66, and BIGb0170 were selected for further analyses. We have included statistics for Figure 1B-C and Figure S1A-G with details mentioned in the figure legend.

(3) Although the behavioral evidence that srd-12 (snif-1) gene encodes a receptor for isoamyl alcohol is compelling, it does not meet the standard for showing that it is an olfactory receptor in C. elegans. To show it is indeed a likely receptor one or more of the following should be done:

(a) Calcium imaging of AWC neurons in response to isoamyl alcohol in the receptor mutant with the expectation that the response would be reduced or abolished in the mutant compared to wildtype.

(b)"A receptor swap" experiment where the SRD-12 (SNIF-1) receptor is expressed in AWB repulsive neuron in SRD-12 (SNIF-1) receptor mutant background with the expectation that with receptor swap C. elegans will now be repulsed from isoamyl alcohol in chemotaxis assays (experiment from Sengupta et al., 1996 odr-10 paper).

Thanks for all your comments and suggestions. While the lab currently does not have the necessary expertise to conduct calcium imaging of neurons, we have performed additional experiments to confirm the requirements of AWC neurons for SNIF-1 function. We generated transgenic worms with extrachromosomal array expressing snif-1 under (a) AWC-specific promoter, odr-1, and (b) AWB-specific promoter, str-1. As shown in new panel 6H in the revised manuscript and Author response image 1, we found that overexpression of snif-1 in AWC neurons completely rescues the chemotaxis defect of snif-1 mutant (referred at VSL2401), whereas upon the “receptor swap" in AWB neurons IAA is sensed as a repellent.

(A) Chemotaxis index (CI) of WT, VSL2401, VSL2401 [AWCp::snif-1] and VSL2401 [AWBp::snif-1] worms to IAA at 1:1000 dilution.

Significant differences are indicated as **** P ≤ 0.0001 determined by one-way ANOVA followed by post hoc Dunnett’s multiple comparison test. Error bars indicate SEM (n≥15).

(4) The authors conclude that C. elegans cannot detect leucine in chemotaxis assays. It is important to add the method for how leucine chemotaxis assay was done in order to interpret these results. Because leucine is not volatile if leucine is put on the plates immediately before the worms are added (as in a traditional odor chemotaxis assay), there is no leucine gradient for the worm to detect. It would be good to put leucine on the plate several hours before worms are introduced so worms have the possibility to be able to detect the gradient of leucine (for example, see Wakabayashi et al., 2009).

Previously, the chemotaxis assays with leucine were performed like traditional odor chemotaxis assays. We also performed chemotaxis assay as detailed in Shingai et al 2005[3]. Leucine was spotted on the assay plates 5 hours prior to the introduction of worms on the plates. As shown in new panel S1H in the revised manuscript, wild-type worms do not show response to leucine in the modified chemotaxis assay.

We have included the experimental details for leucine chemotaxis assays in the revised manuscript.

(5) The bacterial preference assay entitled "odor-only assay" is a misleading name. In the assay, C. elegans is exposed to both volatile chemicals (odors) and non-volatile chemicals because the bacteria are grown on the assay plate for 12 hours before the worms are introduced to the assay plate. In that time, the bacteria is likely releasing non-volatile metabolites into the plate which may affect the worm's preference. A true odor-only assay would have the bacteria on the lid and the worms on the plate.

The ‘odor-only’ diet preference assay does not allow for non-volatile chemicals to reach worms. We achieved this by using tripartite dishes where the compartments containing worms and bacterial odors are separated by polystyrene barriers. At the time of the assay, worms were spotted in a separate compartment from that of bacteria (as shown in schematic 1A). The soluble metabolites released by the bacteria during their growth will accumulate in the agar within the bacterial compartment alone such that worms only encounter the volatile metabolites produced by bacteria wafting past the polystyrene barrier.

(6) The findings of the study should be discussed more in the context of prior literature. For example, AWC neurons have been previously shown to be involved in bacterial preference (Harris et al., 2014; Worthy et al., 2018). In addition, CeMbio bacterial strains (the strains examined in this study) have been previously shown to release isoamyl alcohol (Chai et al. 2024).

Thanks for the suggestion. We have modified the Discussion section to discuss the study in the light of relevant prior literature.

Reviewer #2 (Public review):

Summary:

Siddiqui et al. show that C. elegans prefers certain bacterial strains that have been supplemented with the essential amino acid (EEA) leucine. They convincingly show that some leucine enriched bacteria stimulate the production of isoamyl alcohol (IAA). IAA is an attractive odorant that is sensed by the AWC. The authors an identify a receptor, SRD-12 (SNIF-1), that is expressed in the AWC chemosensory neurons and is required for chemotaxis to IAA. The authors propose that IAA is a predominant olfactory cue that determines diet preference in C. elegans. Since leucine is an EAA, the authors propose that worm IAA sensing allows the animal provides a proxy mechanism to identify EAA rich diets.

Strengths:

The authors propose IAA as a predominant olfactory cue that determines diet preference in C. elegans providing molecular mechanism underlying diet selection. They show that wild isolates of C. elegans have a strong chemotactic response to IAA indicating that IAA is an ecologically relevant odor for the worm. The paper is well written, and the presented data are convincing and well organized. This is an interesting paper that connects chemotactic response with bacterially produced odors and thus provides an understanding of how animals adapt their foraging behavior through the perception of molecules that may indicate the nutritional value.

Weaknesses:

Major:

While I do like the way the authors frame C. elegans IAA sensing as mechanisms to identify leucine (EAA) rich diets it is not fully clear whether bacterial IAA production is a proxy for bacterial leucine levels.

(1) Can the authors measure leucine (or other EAA) content of the different CeMbio strains? This would substantiate the premise in the way they frame this in the introduction. While the authors convincingly show that leucine supplementation induces IAA production in some strains, it is not clear if there are lower leucine levels in the different in non-preferred strains.

Thanks for your suggestion. Estimating leucine levels in various bacteria will provide useful information, and we hope to do so in future studies.

(2) It is not clear whether the non-preferred bacteria in Figure 1A and 1B have the ability to produce IAA. To substantiate the claim that C. elegans prefers CEent1, JUb66, and BIGb0170 due to their ability to generate IAA from leucine, it would measure IAA levels in non-preferred bacteria (+ and - leucine supplementation). If the authors have these data it would be good to include this.

Thanks for the suggestion. We have included the table indicating the presence or absence of IAA production by all the bacteria under + LEU and – LEU conditions (Table S2). Some of the nonpreferred bacteria indeed produce isoamyl alcohol. However, the abundance of IAA in these strains is significantly less than in the preferred bacteria.

Using the available genomic sequence data, we found that all CeMbio strains encode IlvE-like transaminase enzymes[4]. This suggests that presumably all the bacteria have the metabolic capacity to make alpha-ketoisocaproate (an intermediate in IAA biosynthetic pathway) from leucine. However, the regulation of metabolic flux is likely to be quite complex in various bacteria.

(3) The authors would strengthen their claim if they could show that deletion or silencing ilvE enzyme reduces IAA levels and eliminates the increased preference upon leucine supplementation.

We agree that testing worms’ diet preference for the preferred strains upon ilvE knockout will further strengthen the claim for IAA being crucial for finding leucine-enriched diet. Currently the lab does not have the necessary expertise and standardize protocols to do genetic manipulations for the CeMbio strains.

(4) While the three preferred bacteria possess the ilvE gene, it is not clear whether this enzyme is present in the other non-preferred bacterial strains. As far as I know, the CeMbio strains have been sequenced so it should be easy to determine if the non-preferred bacteria possess the capacity to make IAA. Does the expression of ilvE in e.g. E. coli increase its preference index or are the other genes in the biosynthesis pathway missing?

Thanks for the suggestion. Using the available genomic sequence data, we find that all the bacteria in the CeMbio collection possess IlvE-like transaminase necessary for synthesis of alphaketoisocaproate, key metabolite in leucine turn over as well as precursor for IAA [4]. E. coli has an IlvE encoding gene in its genome [2]. However, we do not find IAA in the headspace of E. coli either with or without leucine supplementation. This indicates either (i) E. coli lacks enzymes for subsequent steps in IAA biosynthesis or (ii) leucine provided under the experimental regime is not sufficient to shift the metabolic flux to IAA production.

Previous studies have suggested that in yeast, the final two steps leading to IAA production are catalyzed by decarboxylase and dehydrogenase enzymes1. The genomic and metabolic flux data available for CeMbio do not describe specific enzymes leading up to IAA synthesis [4].

(5) It is strongly implied that leucine-rich diets are beneficial to the worm. Do the authors have data to show the effect on leucine supplementation on C. elegans healthspan, life-span or broodsize?

Edwards et al. 2015 reported a 15% increase in the lifespan of worms upon 1 mM leucine supplementation [5]. Wang et al 2018 also showed lifespan extension upon 1 mM and 10 mM leucine supplementation. They also reported that while leucine supplementation did not have any effect on brood size, it did make worms more resistant to heat, paraquat, and UV-stress [6]. These studies have been included in the discussion section.

Other comments:

Page 6. Figure 2c. While the authors' conclusions are correct based on AWC expts. it would be good at this stage to include the possibility that odors that enriched in the absence of leucine may be aversive.

Thanks for the comment. We have tested the chemotaxis response of the worms for most of the odors produced by CeMbio strains without leucine supplementation. We did not find any odor that is aversive to worms. However, we cannot completely rule out the possibility that a low abundance of aversive odor in the headspace of the bacteria was missed.

Interestingly, we did identify 2-nonanone, a known repellent, in the headspace of the preferred bacteria upon leucine supplementation. However, the abundance of 2-nonanone in headspace of bacteria is relatively low (less than 1% for CEent1, and JUb66, and ~10% for BIGb0170). This suggests that the relative abundance of odors in an odor bouquet may be a relevant factor in determining worms’ reference.

Page 6. IAA increases 1.2-4 folds upon leucine supplementation. If the authors perform a chemotaxis assay with just IAA with 1-2-4 fold differences do you get the shift in preference index as seen with the bacteria? i.e. is the difference in IAA concentration sufficient to explain the shift in bacterial PI upon leucine supplementation? Other attractants such as Acetoin and isobutanol go up in -Leu conditions.

Thanks for the suggestion. As shown in Figure S2H and S2I, when given a choice between a concentration of IAA (1:1000 dilution) attractive to worms and a 4-fold higher amount of IAA, worms chose the latter. This result suggests that worms can distinguish between relatively small difference in concentrations of IAA.

We agree that the relative abundance of Acetoin and Isobutanol is high in -LEU conditions. The presence of other attractants in - LEU conditions should skew the preference of worms for – LEU bacteria. However, we found that worms prefer + LEU bacteria (Figure 1B), suggesting that the abundance of IAA mainly influences the diet preference of the worms.

Page 14-15. The authors identify a putative IAA receptor based on expression studies. I compliment the authors for isolating two CRISPR deletion alleles. They show that the srd-12 (snif-1) mutants have obvious defects in IAA chemotaxis. Very few ligand-odorant receptors combinations have been identified so this is an important discovery. CenGen data indicate that srd-12 (snif-1) is expressed in a limited set of neurons. Did the authors generate a reporter to show the expression of srd-12 (snif-1)? This is a simple experiment that would add to the characterization of the SRD-12 (SNIF-1) receptor. Rescue experiments would be nice even though the authors have independent alleles. To truly claim that SRD-12 (SNIF-1) is the ligand for IAA and activates the AWC neurons would require GCamp experiments in the AWC neuron or heterologous expression system. I understand that GCamp imaging might not be part of the regular arsenal of the lab but it would be a great addition (even in collaboration with one of the many labs that do this regularly). Comparing AWC activity using GCaMP in response IAA-producing bacteria with high leucine levels in both wild-type and SRD-12 (SNIF-1) deficient backgrounds, would further support their narrative. I leave that to the authors.

Thanks for your comments and suggestions. To address this comment, we rescued snif-1 mutant (referred as VSL2401) with extrachromosomal array expressing snif-1 under AWC-specific promoter as well as its native promoter. As shown in Figure 6H and Author response image 2, we find that both transgenic lines show a complete rescue of chemotaxis response to isoamyl alcohol. To find where snif-1 is expressed, we generated a transgenic line of worms expressing GFP under snif-1 promoter, and mCherry under odr-1 promoter (to mark AWC neurons). As shown in Figure 6I, we found that snif-1 is expressed faintly in many neurons, with strong expression in one of the two AWC neurons marked by odr-1::mCherry. This result suggests that SNIF-1 is expressed in AWC neuron.

We hope to perform GCaMP assay and further characterization of SNIF-1 in the future.

Chemotaxis index (CI) of WT, VSL2401, VSL2401 [AWCp:: snif-1] and VSL2401 [snif-1p::snif-1] worms to IAA at 1:1000 dilution.

Significant differences are indicated as **** P ≤ 0.0001 determined by one-way ANOVA followed by post hoc Dunnett’s multiple comparison test. Error bars indicate SEM (n≥15).

Minor:

Page 4 "These results suggested that worms can forage for diets enriched in specific EAA, leucine...." More precise at this stage would be to state " These results indicated that worms can forage for diets supplemented with specific EAA...".

We have changed the statement in the revised manuscript.

Page 5."these findings suggested that worms not only rely on odors to choose between two bacteria but also to find leucine enriched bacteria" This statement is not clear to me and doesn't follow the data in Fig. S2. Preferred diets in odorant assays are the IAA producing strains.

Thanks for your comment. We have revised the manuscript to make it clear. “Altogether, these findings suggested that worms rely on odors to distinguish different bacteria and find leucineenriched bacteria”. This statement concludes all the data shown in Figure 1 and Figure S1.

Page 5. Figure S2A provides nice and useful data that can be part of the main Figure 1.

Thanks for the comment. We have incorporated the data from Figure S2A to main Figure 1.

Reviewer #3 (Public review):

Summary:

The authors first tested whether EAA supplementation increases olfactory preference for bacterial food for a variety of bacterial strains. Of the EAAs, they found only leucine supplementation increased olfactory preference (within a bacterial strain), and only for 3 of the bacterial strains tested. Leucine itself was not found to be intrinsically attractive.

They determined that leucine supplementation increases isoamyl alcohol (IAA) production in the 3 preferred bacterial strains. They identify the biochemical pathway that catabolizes leucine to IAA, showing that a required enzyme for this pathway is upregulated upon supplementation.

Consistent with earlier studies, they find that AWC olfactory neuron is primarily responsible for increased preference for IAA-producing bacteria.

Testing volatile compounds produced by bacteria and identified by GC/MS, and identified several as attractive, most of them require AWC for the full effect. Adaptation assays were used to show that odorant levels produced by bacterial lawns were sufficient to induce olfactory adaptation, and adaptation to IAA reduced chemotaxis to leucine-supplemented lawns. They then showed that IAA attractiveness is conserved across wild strains, while other compounds are more variable, suggesting IAA is a principal foraging cue.

Finally, using the CeNGEN database, they developed a list of candidate IAA receptors. Using behavioral tests, they show that mutation of srd-12 (snif-1) greatly impairs IAA chemotaxis without affecting locomotion or attraction to another AWC-sensed odor, PEA.

Comments

This study will be of great interest in the field of C. elegans behavior, chemical senses and chemical ecology, and understanding of the sensory biology of foraging.

Strengths:

The identification of a receptor for IAA is an excellent finding. The combination of microbial metabolic chemistry and the use of natural bacteria and nematode strains makes an extremely compelling case for the ecological and adaptive relevance of the findings.

Weaknesses:

AWC receives synaptic input from other chemosensory neurons, and thus could potentially mediate navigation behaviors to compounds detected in whole or in part by those neurons. Language concluding detection by AWC should be moderated (e.g. p9 "worms sense an extensive repertoire...predominantly using AWC") unless it has been demonstrated.

Thanks for your comment. We have modified the manuscript to incorporate the suggestion.

srd-12 (snif-1) is not exclusively expressed in AWC. Normally, cell-specific rescue or knockdown would be used to demonstrate function in a specific cell. The authors should provide such a demonstration or explain why they are confident srd-12 (snif-1) acts in AWC.

Thanks for the comment. We have performed AWC-specific rescue of snif-1 in mutant worms. As shown in Figure 6H, we found that AWC neurons specific rescue completely recovered the chemotaxis defect of the snif-1 mutant (referred as VSL2401) for IAA. In addition, snif-1 is expressed in one of the AWC neurons.

A comparison of AWC's physiological responses between WT and srd-12 (snif-1), preferably in an unc13 background, would be nice. Even further, the expression of srd-12 (snif-1) in a different neuron type and showing that it confers responsiveness to IAA (in this case, inhibition) would be very convincing.

Thanks for the suggestion. We have performed a receptor swap experiment, where snif-1 is misexpressed in AWB neurons. We find that these worms show slight but significant repulsion to IAA compared to WT and snif-1 mutant worms (Author response image 1).

Recommendations for the authors:

Reviewing Editor:

Please consider all of the reviewer comments. In particular, as noted in the individual reviews, the strength of the evidence would be bolstered by additional experiments to demonstrate that the iLvE enzyme affects IAA levels in the preferred bacteria. The reviewers note that the authors haven't shown that IAA production is a reflection of leucine content. Are the non-preferred bacteria low on leucine or lack iLvE or IAA synthesis pathways? Further, more direct evidence that SRD-12 (SNIF-1) is in fact the primary IAA receptor would further strengthen the study. The authors should also be aware that geographic distance for wild isolate C. elegans may not directly correlate with phylogenetic distance. This should be assessed/discussed for the strains used.

Thanks for the suggestions. Some of these have been addressed in response to reviewers. Thanks for your comments about possible disconnect between geographical and phylogenetic distances amongst natural isolates used here.

By analyzing the phylogenetic tree generated using neighbor-joining algorithm available at CaeNDR database, we found that QX1211 and JU3226 are phylogenetically close, but the remaining isolates fall under different clades separated by long phylogenetic distances [7,8].

Reviewer #1 (Recommendations for the authors):

(1) In the first sentence of the third paragraph of the introduction, C. elegans are described as "soildwelling." Although C. elegans has been described as soil-dwelling in the past, current research indicates they are most often found on rotten fruit, compost heaps and other bacterial-rich environments, not soil. "All Caenorhabditis species are colonizers of nutrient- and bacteria-rich substrates and none of them is a true soil nematode." from Kiontke, K. and Sudhaus, W. Ecology of Caenorhabditis species (WormBook).

Your specific comment about C. elegans’ habitat is well received. However, in that sentence we are referring to the chemosensory system of soil-dwelling animals in general, and not particularly C. elegans.

(2) Figure 3K, the model would be clearer if leucine-rich diet -> volatile chemicals ->AWC (instead of leucine-rich diet -> AWC <- volatile chemicals). The leucine-rich diet results in the production of volatile chemicals which are detected by AWC.

We have modified the figure to make it clearer.

(3) Figure 4 - it would help to include a table summarizing the volatile chemicals that each bacteria releases. Then the reader could more easily evaluate whether the adaptation to each specific odor is consistent with the change in preference for the specific bacteria based on what it releases in its headspace. In addition, Figure 4 would help to clarify whether bacteria in these experiments were cultured with or without leucine supplementation.

Table S2 summarizes the odors released by all the bacteria under + LEU and – LEU conditions.

In Figure 4, adaptation was performed by odors of bacteria when cultured under leucineunsupplemented conditions.

Reviewer #2 (Recommendations for the authors):

Page 9. Previous studies e.g. Bargmann Hartwieg and Horvitz have shown IAA is sensed by the AWC. Would be good to cite appropriately.

Thanks for the comment. The reference has been cited at p9 and p16.

References:

(1) Yuan, J., Mishra, P., and Ching, C.B. (2017). Engineering the leucine biosynthetic pathway for isoamyl alcohol overproduction in Saccharomyces cerevisiae. Journal of Industrial Microbiology and Biotechnology 44, 107-117. 10.1007/s10295-016-1855-2 %J Journal of Industrial Microbiology and Biotechnology.

(2) Kanehisa, M., Furumichi, M., Sato, Y., Matsuura, Y., and Ishiguro-Watanabe, M. (2025). KEGG: biological systems database as a model of the real world. Nucleic Acids Res 53, D672-d677. 10.1093/nar/gkae909.

(3) Shingai, R., Wakabayashi, T., Sakata, K., and Matsuura, T. (2005). Chemotaxis of Caenorhabditis elegans during simultaneous presentation of two water-soluble attractants, llysine and chloride ions. Comparative biochemistry and physiology. Part A, Molecular & integrative physiology 142, 308-317. 10.1016/j.cbpa.2005.07.010.

(4) Dirksen, P., Assié, A., Zimmermann, J., Zhang, F., Tietje, A.M., Marsh, S.A., Félix, M.A., Shapira, M., Kaleta, C., Schulenburg, H., and Samuel, B.S. (2020). CeMbio - The Caenorhabditis elegans Microbiome Resource. G3 (Bethesda, Md.) 10, 3025-3039. 10.1534/g3.120.401309.

(5) Edwards, C., Canfield, J., Copes, N., Brito, A., Rehan, M., Lipps, D., Brunquell, J., Westerheide, S.D., and Bradshaw, P.C. (2015). Mechanisms of amino acid-mediated lifespan extension in Caenorhabditis elegans. BMC genetics 16, 8. 10.1186/s12863-015-0167-2.

(6) Wang, H., Wang, J., Zhang, Z.J.J.o.F., and Research, N. (2018). Leucine Exerts Lifespan Extension and Improvement in Three Types of Stress Resistance (Thermotolerance, AntiOxidation and Anti-UV Irradiation) in C. elegans. 6, 665-673.

(7) Crombie, T.A., McKeown, R., Moya, N.D., Evans, Kathryn S., Widmayer, Samuel J., LaGrassa, V., Roman, N., Tursunova, O., Zhang, G., Gibson, Sophia B., et al. (2023). CaeNDR, the Caenorhabditis Natural Diversity Resource. Nucleic Acids Research 52, D850-D858. 10.1093/nar/gkad887 %J Nucleic Acids Research.

(8) Cook, D.E., Zdraljevic, S., Roberts, J.P., and Andersen, E.C. (2017). CeNDR, the Caenorhabditis elegans natural diversity resource. Nucleic Acids Res 45, D650-d657. 10.1093/nar/gkw893.

Version 2

Abstract

Acquisition of essential nutrients through diet is crucial for the survival of animals. Dietary odors might enable animals to forage for nutrient-rich diets. We asked if Caenorhabditis elegans, a bacterivorous nematode, uses olfactory cues to forage for essential amino acid-rich (EAA) diets. Using the native microbiome of C. elegans, we show that worms rely on olfaction to select leucine (EAA)-enriched bacteria. Using gas chromatography, we find that leucine-enriched bacteria produce isoamyl alcohol (IAA) odor in the highest abundance. Prior adaptation of worms to IAA diminishes the diet preference of worms. Several wild isolates of C. elegans display robust responses to IAA emphasizing its ecological relevance. We find that foraging for a leucine-enriched diet is mediated via the AWC olfactory neurons. Finally, we identify SNIF-1 G protein-coupled receptor in AWC neurons as a receptor for IAA and a mediator of dietary decisions in worms. Our study identifies a receptor-ligand module underpinning foraging behavior in C. elegans.

Introduction

Amino acids are central to metabolism, not only as building blocks for proteins but also as signaling molecules. During the course of evolution, animals have lost the ability to synthesize half of the proteogenic amino acids, making dietary intake of such amino acids indispensable (Gietzen and Rogers, 2006). Leucine (Leu), isoleucine (Ile), valine (Val), histidine (His), lysine (Lys), tryptophan (Trp), phenylalanine (Phe), methionine (Met), threonine (Thr), and arginine (Arg) constitute essential amino acids (EAAs) (Tomlinson et al., 2011). Several studies have investigated the mechanism of amino acid surveillance by cells and tissues. Leucine, glutamine, and arginine have been shown to activate the mTOR pathway and control protein translation and cell growth (Sancak et al., 2008, Hara et al., 1998, van der Vos and Coffer, 2012, Bauchart-Thevret et al., 2010, Gauthier-Coles et al., 2021, Bar-Peled and Sabatini, 2014, González and Hall, 2017). Altered EAA levels influence the longevity, behavior, and health of animals (Austad et al., 2024, Minussi et al., 2023, Enomoto et al., 2013). Increased levels of circulating branched-chain amino acids (BCAAs) are associated with an increased risk of obesity and diabetes in humans (Tai et al., 2010, Wang et al., 2011). Restricted intake of tryptophan and methionine extends lifespan in rats, while a tryptophan-deficient diet reduces the weight of organs (De Marte and Enesco, 1986, Miller et al., 2005). Methionine restriction is reported to extend the lifespan of Drosophila (Lee et al., 2014). Central administration of a BCAA, leucine, reduces food intake in rats (Cota et al., 2006). An extended lifespan upon diet restriction in silkworms is linked to enhanced utilization of EAAs (Wang et al., 2023). However, it is not clear if animals have a way to modulate EAA-specific dietary intake. Do animals use specific sensing mechanisms to find an EAA-enriched diet? Several studies on feeding behaviors suggest that both invertebrates and vertebrates select nutrient-rich diets. For instance, rats prefer a protein-rich diet over a carbohydrate-rich diet (Makarios-Lahham et al., 2004). Many animals display mechanisms to choose for or reject food with specific EAAs. Sparrows consume a combination of different diets to get adequate levels of various EAAs, while gastropods reject diets deficient in methionine (Delaney and Gelperin, 1986, Gietzen, 1993, Murphy and Pearcy, 1993). When maintained on an EAA-restricted diet, flies compensate by increasing feeding (Juricic et al., 2019). Kittens prefer diets rich in methionine and threonine, while rats select for a leucine-enriched diet (Rogers et al., 2004). Despite evidence that organisms seek EAA, specific molecular mechanisms for search behavior are not well understood.

Chemoperception, taste and olfaction via G-protein coupled receptors (GPCRs) allow animals to quickly assess their diet. EAAs evoke taste perceptions such as sweet (Thr, His, Leu, Phe, Trp) and bitter (Arg, Ile, Lys, Met), while non-EAAs can elicit umami (Glu) and sour (Asp) taste (Shallenberger, 1993, Schiffman et al., 1981, Zhao et al., 2016). Amino acids that elicit sweet responses in humans, like threonine, are found to be attractive to rodents as well (Bachmanov et al., 2016, Yoshida and Saito, 1969). GPCRs are implicated in sensing different classes of amino acids and regulate foraging behavior in animals (Wellendorph et al., 2005b, Kuang et al., 2003, Nelson et al., 2002, Conigrave et al., 2000, Conigrave and Hampson, 2006). Metabotropic glutamate receptors (mGlu) sense glutamate, while the extracellular calcium-sensing receptor senses aliphatic, aromatic, and polar amino acids (Conigrave et al., 2000, Nakanishi, 1992). Invertebrates like Caenorhabditis elegans and Drosophila melanogaster also possess GPCR-based amino acid-sensing mechanisms (Miguel-Aliaga, 2012). C. elegans senses glutamate using mGlu type receptors, Mgl-1/2/3, while D. melanogaster utilizes a cationic amino-acid transporter, Dm slif, to sense arginine (Dillon et al., 2015, Colombani et al., 2003). The goldfish 5.24, an odorant receptor, responds to arginine, while its human homolog, GPRC6A, responds to basic as well as aliphatic amino acids (Speca et al., 1999, Wellendorph et al., 2005a). Olfactory GPCRs are known to sense various odors, including alcohol, esters, ketones, etc., and many of these are products of EAA catabolism. It is not known if animals rely on EAA-derived odors to find an EAA-rich diet.

The presence of an elaborate chemosensory system in soil-dwelling animals suggests the usage of olfactory cues for foraging. C. elegans, a bacterivorous nematode, relies heavily on its simple but effective chemosensory system composed of 302 neurons for foraging, although specific cues regulating this phenomenon are unknown (Sawin et al., 2000, Rhoades et al., 2019). Olfactory neurons of C. elegans express as many as 335 GPCRs (Hammarlund et al., 2018), however, the odors stimulating them remain largely unknown. Like other invertebrates, C. elegans requires ten EAAs, of these, leucine, tryptophan, valine, arginine, and lysine extend their lifespan when supplemented at low concentration (Edwards et al., 2015). Does C. elegans rely on olfactory cues to sense diets with high(er) levels of EAAs? It is conceivable since some bacteria possess biosynthetic pathways to convert amino acids into odors, such as dimethyl sulfide, indole, isoamyl acetate, phenylethyl alcohol, and benzyl alcohol from methionine, tryptophan, leucine, and phenyl alanine respectively (Weisskopf et al., 2021). C. elegans can sense hundreds of odors representing extensive structural diversity (Bargmann et al., 1993). Do worms utilize some of them as foraging signals to find nutrient-enriched bacteria?

We hypothesized that certain odors produced by the microbiome of C. elegans are olfactory signals for EAA-rich bacteria and drive worms’ foraging behavior. To test this hypothesis, we analyzed the preference of worms for natural microbiota species supplemented with individual EAAs. We found that odors drive worms’ feeding preference in favor of leucine-supplemented diets. By carefully analyzing the odor bouquet produced by natural microbiota upon leucine supplementation, we identified isoamyl alcohol (IAA) as the most abundant odor in preferred diets. We show that IAA is produced from leucine using the Ehrlich degradation pathway in bacteria. IAA is an ecologically relevant odor and regulates the preference of worms for a leucine-supplemented diet. Finally, we identified SNIF-1, a GPCR expressed in AWC neurons, as the receptor for IAA. Taken together, SNIF-1 regulates the dietary preference of worms to IAA-producing bacteria and thereby mediates the foraging behavior of C. elegans to leucine-enriched diets. Thus, IAA produced by bacteria is a dietary quality code for leucine-enriched bacteria.

Results

Microbial odors drive the preference of C. elegans for leucine-enriched diets

To understand if C. elegans used olfactory signals to distinguish between standard and EAA-enriched diets, we performed ‘odor-only’ diet preference assays using natural microbiota of C. elegans, CeMbio, and laboratory food, Escherichia coli OP50. CeMbio is a collection of 12 bacteria that best reflects the metabolic complexity of the microbes in the natural habitat of worms (see Table S1). N2 or wild type (WT) worms were allowed to choose between odors from each strain grown on separate sections, EAA supplemented (+ EAA) or unsupplemented (-EAA), of a tripartite plate (schematic in Figure 1A). Of the ten EAAs tested, WT worms showed a preference for all the bacteria supplemented with leucine-supplemented (+ LEU) over unsupplemented (-LEU) except for Chryseobacterium scophthalmum (JUb44) and Escherichia coli OP50 (Figure 1B). However, worms could not distinguish between supplemented and unsupplemented bacteria for the remaining nine EAAs for most of the diets (Figures 1C, 1D and S1A-S1G). We found that worms did not respond to leucine at three different concentrations in a modified chemotaxis assay, suggesting that worms are incapable of sensing leucine directly (Figure S1H). These results suggested that worms can forage for diets supplemented with specific EAA, leucine, using odors produced by some bacteria.

C. elegans relies on odors to select leucine-enriched bacteria.

(A) Schematic representation of ‘odor-only’ diet preference assay. Preference index (PI) of WT worms in an ‘odor-only’ diet preference assay (indicated as ) for individual diet supplemented with (B) 5 mM leucine (+ LEU), (C) 5 mM isoleucine (+ ILE), and (D) 5 mM valine (+ VAL). Significant differences are indicated as * P ≤ 0.05, ** P ≤ 0.01, and **** P ≤ 0.0001 determined by one-sample t-test. (E) Schematic representation of diet preference assay between individual CeMbio bacteria and E. coli OP50. (F) Preference index (PI) of WT worms in a diet preference assay between individual CeMbio bacteria and E. coli OP50. Also, see movie S1. Error bars indicate SEM (n≥15).

We next asked whether worms prefer their native microbiome over laboratory food E. coli OP50. We performed ‘odor-only’ diet preference assays to test if worms can distinguish between CeMbio and E. coli OP50 using odors (schematic in Figure 1E). We found that the preference indices (PI) of worms for Enterobacter hormaechei (CEent1), Lelliottia amnigena (JUb66), and Sphingobacterium multivorum (BIGb0170) were positive (preferred diets) while neutral or negative for other strains such as Pseudomonas lurida MYb11 (Figure 1F). Altogether, these findings suggested that worms rely on odors to distinguish various bacteria and find leucine-enriched bacteria.

Isoamyl alcohol odor is a signature for a leucine-enriched diet

We hypothesized that bacterial odor(s) enriched upon leucine supplementation are likely to be drivers of foraging behavior in worms. We identified all the odors in the headspace of individual bacteria using solid-phase microextraction followed by gas chromatography-mass spectrometry (GC-MS/MS). By examining levels of individual odors in leucine-supplemented and unsupplemented bacteria, we found that several odors were present in the headspace of bacteria (Figures 2A, 2B, and S2A-S2D, also see table S2). However, isoamyl alcohol (IAA) was the most abundant and the only shared odor in the worms’ preferred diets, CEent1, JUb66, and BIGb0170 (Figures 2A-2C and S2A-S2D). We found that under standard conditions, IAA constituted 40-70% of the headspace of preferred bacteria and remarkably increased up to 90% upon leucine supplementation (Figure 2C). We also quantified leucine-dependent increase in IAA levels in the headspace of preferred bacteria (Figure S3E, see methods). We found that the abundance of IAA significantly increased in CEent1, JUb66, and BIGb0170 up to 4 folds upon leucine supplementation (Figures 2D, S2F and S2G). To test if 4-fold increase in IAA concentration is sufficient to explain worms’ preference to leucine-supplemented bacteria, we performed a chemotaxis assay where worms were given a choice between 4-fold and 1-fold concentration of IAA (schematic in Figure S2H). We found that worms prefer increased concentration of IAA (Figure S2I). This suggested that worms can sense the gradient of IAA in chemically complex scenario.

Leucine supplementation boosts isoamyl alcohol levels via Ehrlich degradation pathway.

GC-MS/MS profile of odors produced by CEent1 under (A) leucine-supplemented and (B) leucine-unsupplemented conditions. Unmarked peaks represent masses contributed by fiber or media alone (n≥3). (C) Heat map representing the relative abundance of various odors in the headspace of CEent1, JUb66, and BIGb0170 with and without leucine supplementation. (D) Absolute abundance of IAA (in moles) produced by CEent1 under leucine-supplemented and unsupplemented conditions. * P ≤ 0.05 as determined by a two-tailed unpaired t-test. (E) Schematic of the Ehrlich degradation pathway. (F) Fold change of transcript levels of ilvE under leucine-supplemented over unsupplemented conditions for CEent1. Error bars indicate SEM (n≥3).

Microbes can produce short-chain alcohols from carbon sources as well as branched-chain amino acids (Weisskopf et al., 2021). Leucine can be catabolized to IAA via the Ehrlich degradation pathway in three enzymatic steps. The first step is catalyzed by IlvE, a transaminase, which converts leucine to α-Ketoisocaproate (schematic in Figure 2E). The ilvE gene is present in the genome of CEent1, JUb66, and BIGb0170 (Dirksen et al., 2020). We predicted that leucine supplementation would result in the upregulation of ilvE in a substrate-dependent manner. Using qRT-PCR, we found that transcript levels of ilvE were 2-fold higher in leucine-supplemented CEent1 over unsupplemented (Figure 2F). This indicated that the Ehrlich degradation pathway is functional in IAA-producing bacteria and further stimulated in response to leucine. Altogether, these findings show that IAA produced via Ehrlich degradation is a signature for a leucine-enriched bacteria.

AWC odor sensory neurons facilitate the diet preference of C. elegans for leucine-enriched diets

Worms use two pairs of odor sensory neurons, AWA and AWC, to sense attractive olfactory cues (Bargmann, 2006, Bargmann et al., 1993, Troemel et al., 1997b). To identify the odor sensory neurons that mediate preference for leucine-enriched diets, we studied the response of odr-7 worms, which do not have functional AWA neurons, and AWC ablation (-) worms (Beverly et al., 2011, Sengupta et al., 1994). We found that the preference of worms for leucine-supplemented diets was dramatically lost in AWC(-) worms but was retained in odr-7 worms (Figures 3A-3C). This suggested that AWC neurons play a crucial role in sensing odors from leucine-enriched diets. We also tested the role of these neurons in mediating diet preference between preferred CeMbio bacteria and E. coli OP50. We found that AWC(-) worms showed reduced preference for CEent1, JUb66, and BIGb0170 over E. coli, while odr-7 mutants had slightly diminished preference for CEent1 and BIGb0170 over E. coli (Figures S3A-S3C). This suggested AWC, largely, and AWA odor sensory neurons contribute to the diet preference of worms for specific microbes in their natural habitat (Figure S3D). The finding that diet preference entails two pairs of neurons suggests that multiple odors constitute a foraging signal enabling worms to find diet bacteria.

Odors sensed through AWC neurons mediate the diet preference of C. elegans.

Preference index (PI) of WT, odr-7, and AWC(-) worms for (A) CEent1, (B) JUb66, and (C) BIGb0170 supplemented with leucine over unsupplemented conditions. symbol indicates ‘odor-only’ preference assays. Chemotaxis index (CI) of WT, odr-7, and AWC(-) worms for (D) isoamyl alcohol (IAA), (E) acetoin (ACE), (F) isovalerate (ISV), (G) isobutanol (ISB), (H) phenylethyl alcohol (PEA), (I) butyl acetate (BA), and (J) methyl isovalerate (MIV) (K) Summary schematic representing the role of AWA and AWC odor sensory neurons in ‘odor-only’ diet preference for leucine-enriched diets and chemotaxis assays. Significant differences are indicated as * P ≤ 0.05, ** P ≤ 0.01, *** P ≤ 0.001, and **** P ≤ 0.0001 determined by one-way ANOVA followed by post hoc Dunnett’s multiple comparison test. Error bars indicate SEM (n=15).

To identify the constituents of the foraging signal, we sampled the headspace of each CeMbio bacterium. Each bacterium produced a unique bouquet of odors composed of chemically diverse molecules, including alcohols, aldehydes, ketones, esters, and carboxylic acids (Table S2). To identify foraging cues amongst odors in the headspace of preferred bacteria, we examined the response of WT worms to ten different odors in chemotaxis assays (schematic in Figure S4A). We found that WT worms were attracted to IAA, acetoin (ACE), isovalerate (ISV), isobutanol (ISB), and phenylethyl alcohol (PEA) in a dose-dependent manner (Figures S4B-S4F), of which IAA has been reported to be a strong attractant. Worms showed attraction to butyl acetate (BA) and methyl isovalerate (MIV) only at specific concentrations but did not display a clear dose-response (Figures S4G and S4H). Worms showed no response to 2,3-butanediol (BDL), isoamyl acetate (ISA), and indole (IND) (Figures S4I-S4K). We would like to note that this is the first report of ACE and MIV as attractants for C. elegans. These results show that the headspace of preferred bacteria contains multiple attractive odors for C. elegans.

To identify neurons responsible for sensing attractive odors produced by preferred bacteria, we used the odr-7 and AWC(-) worms in chemotaxis assays for each of the seven attractive odors. We found that AWC(-) worms had severely diminished responses to IAA, ISV, ISB, PEA, BA, and MIV (Figures 3D and 3F-3J). ACE, on the other hand, was sensed by AWA neurons as odr-7 showed reduced response to acetoin (Figure 3E). Taken together, our findings revealed that worms sense an extensive repertoire of chemically diverse odors produced by preferred bacteria as attractants, the majority of them using AWC neurons (summary in Figure 3K).

Isoamyl alcohol drives foraging behavior in C. elegans

To identify the most relevant foraging signal for C. elegans, we took advantage of the well-established odor adaptation assay using each of the attractive odors in its diet (schematic in Figure 4A). Prolonged exposure of C. elegans to an odorant is known to result in the loss or reduction of the ability of worms to sense the same odorant in a subsequent exposure (Colbert and Bargmann, 1995). We performed a modified adaptation assay where worms were exposed to a bouquet of odors from a bacterial lawn (-LEU), followed by testing their chemotaxis response to individual attractive odors. Worms adapted to odors from CEent1 lawns had dramatically reduced responses to IAA, isobutanol, and phenylethyl alcohol compared to naïve worms (Figures 4B, 4E, and 4F). Worms adapted to odors from JUb66 and BIGb0170 lawns also had dramatically reduced response to IAA (Figure 4B). While adaptation to bacterial odors did not influence worms’ response to ACE, ISV, BA, and MIV (Figure 4C, 4D, 4G and 4H). These findings suggested that the odor bouquets of preferred bacteria have levels of IAA high enough to modulate chemoperception in C. elegans.

Isoamyl alcohol regulates the diet preference of worms.

(A) Schematic representation of the adaptation regimen followed by chemotaxis assays. Chemotaxis index (CI) for naïve worms and worms adapted with CEent1, JUb66, or BIGb0170 odors to (B) isoamyl alcohol (IAA), (C) acetoin (ACE), (D) isovalerate (ISV), (E) isobutanol (ISB) (F) phenylethyl alcohol (PEA), (G) butyl acetate (BA), and (H) methyl isovalerate (MIV). Dark gray bars indicate naïve worms, and light gray bars indicate worms adapted to bacterial odors. Significant differences are indicated as * P ≤ 0.05, ** P ≤ 0.01, *** P ≤ 0.001, and **** P ≤ 0.0001 determined by one-way ANOVA followed by post hoc Dunnett’s multiple comparison test. Error bars indicate SEM (n=15).

If IAA was the most relevant foraging signal, we predicted that prior adaptation to IAA would diminish the diet preference of C. elegans. To test this, we adapted worms to IAA odor and tested their diet preference (schematic in Figure S5A). IAA-adapted worms showed a reduced chemotaxis response to IAA, indicating that the adaptation regimen is robust (Figure S5B). We found that IAA-adapted worms displayed a reduced preference for CEent1, JUb66, or BIGb0170 over E. coli OP50 compared to naïve worms (Figures S5C-S5E). These results indicated that IAA was the predominant olfactory cue that determines the diet preference of C. elegans.

IAA is an ecologically relevant odor for C. elegans

If an odor is relevant as a foraging cue in nature, the ability to sense it must be under positive selection in wild C. elegans populations. We anticipated that most wild isolates of C. elegans will have a robust response to IAA but not to other attractants reported in this study. To test this hypothesis, we studied the response to IAA in nine phylogenetically distinct wild isolates of C. elegans collected from geographically distinct regions (Figure 5A and Table S1) (Andersen et al., 2012, Cook et al., 2017). Additionally, we tested their chemotaxis response to diacetyl, a known food cue, and two other attractants (PEA and ACE) identified in this study. We found that all the strains displayed robust response to IAA comparable to WT worms (Figure 5B). The chemotaxis response of these strains to diacetyl was also quite robust (Figure 5C). However, the chemotaxis response to PEA and acetoin were variable within and across wild isolates suggesting they may not be crucial for foraging (Figures 5D and 5E). The robust response of wild isolates to IAA supports the notion that the ability of C. elegans to sense IAA has been selected during evolution. These findings suggest that IAA is an ecologically relevant foraging cue for C. elegans.

Robust chemoperception of isoamyl alcohol in wild isolates of C. elegans.

(A) World map representing the distinct geographical locations (source) of wild isolates of C. elegans used in this study. Chemotaxis index (CI) of WT and wild isolates of worms for (B) IAA, (C) diacetyl (DA), (D) phenylethyl alcohol (PEA), and (E) acetoin (ACE). Significant differences are indicated as * P ≤ 0.05, ** P ≤ 0.01, *** P ≤ 0.001, and **** P ≤ 0.0001 determined by one-way ANOVA followed by post hoc Dunnett’s multiple comparison test. Error bars indicate SEM (n=15).

SNIF-1 senses isoamyl alcohol and facilitates foraging in C. elegans

C. elegans genome encodes for ∼1300 putative GPCRs, many of which are predicted to sense soluble and olfactory cues (Thomas and Robertson, 2008). Since the preference for IAA and leucine-enriched diets is mediated by AWC odor sensory neurons (Figures 3A-3D), we hypothesized that the receptor for IAA is a GPCR expressed in AWC neurons. Indeed, calcium response to IAA has been reported in AWC neurons (Zaslaver et al., 2015, Lin et al., 2023, Bargmann and Horvitz, 1991). Using single-cell RNAseq data for C. elegans neurons, available at CeNGEN (Hammarlund et al., 2018), we listed 18 putative GPCRs expressed in AWCon and AWCoff for further analysis (Figure 6A). We tested response to IAA in 15 strains harboring mutants in one or more of these GPCRs (see Table S1) (Pu et al., 2023b). One of these strains, CHS1169, had a significantly reduced response to IAA compared to WT worms (Figure 6A). Of the six GPCRs encoding genes edited in CHS1169 strain, srd-12 had the highest expression in AWCoff odor sensory neuron (Hammarlund et al., 2018). Using CRISPR/Cas9-based approach, we generated two srd-12 deletion lines, VSL2401, and VSL2402, carrying mutation in srd-12 gene (see methods and Table S1). We recorded the chemotaxis response of WT and mutant worms to IAA (schematic in Figure 6B) and found they were indeed defective in their response to IAA (Figures 6C-6F). By analyzing the locomotion of WT, VSL2401, and VSL2402 worms, in the chemotaxis arena, we determined quantitative differences in their response to IAA. We observed that VSL2401 and VSL2402 strains had speeds comparable to WT worms and traveled distances similar to WT worms suggesting that srd-12 mutants do not have locomotory defects (Figures S6A and S6B). We found that mutation in srd-12 gene resulted insignificantly reduced chemotaxis response to IAA (Figure 6F). However, srd-12 mutants had a comparable response to WT worms for ACE and PEA (Figures S6C and S6D). This result suggested that mutation in srd-12 cause defect in sensing IAA alone. We rename srd-12 gene and encoded protein as snif-1 and sniffing receptor SNIF-1 respectively. As expected, snif-1 mutant strains had diminished preference for CEent1 over E. coli OP50 in the diet preference assay consistent with the idea that IAA is the most relevant foraging signal (Figure 6G).

SNIF-1 GPCR mediates isoamyl alcohol sensing and diet preference in C. elegans.

(A) List of GPCRs that are highly expressed in AWC neurons. Chemotaxis index (CI) of WT and GPCR edited strains to 1:1000 IAA (n≥3). All GPCR mutants of C. elegans used for screening have the CHS designation along with the codes mentioned in the panels. Also, refer to Table S1 for strain information. (B) Schematic for chemotaxis assay plate used for movement track analysis. Movement tracks of 5 animals in a chemotaxis arena with 1:1000 IAA for 15 minutes, for (C) WT, and snif-1 mutants-(D) VSL2401, and (E) VSL2402. Also, refer to video S2. Individual color represents the track for a single worm. (F) Chemotaxis index (CI) in response to IAA (1:1000 dilution) for WT, VSL2401, and VSL2402 worms. (G) Preference index (PI) of WT, VSL2401, and VSL2402 worms for a preferred diet, CEent1, over E. coli OP50. symbol indicates ‘odor-only’ preference assays. (H) Chemotaxis index (CI) in response to IAA (1:1000 dilution) for WT, VSL2401, VSL2401 [snif-1p:: snif-1], VSL2401 [AWCp::snif-1] and VSL2401 [AWBp::snif-1]. Significant differences are indicated as * P ≤ 0.05, ** P ≤ 0.01, and **** P ≤ 0.0001 determined by one-way ANOVA followed by post hoc Dunnett’s multiple comparison test. Error bars indicate SEM (n≥15). (I) Schematic showing the soma and processes of AWC neurons in the amphid region. Representative images of the reporter line expressing GFP under snif-1 promoter, and mCherry under odr-1 promoter (to mark AWC neurons) (n≥30).

To confirm that IAA receptor is indeed functional in AWC neuron, we generated transgenic worms carrying extrachromosomal array expressing snif-1 cDNA under its native promoter (snif-1p:: snif-1) as well as under an AWC-specific promoter (AWCp:: snif-1). We found that transgenic lines expressing SNIF-1 under its native promoter, or AWC-specific promoter completely rescued the chemotaxis response of the snif-1 mutant worms to IAA (Figure 6H). To further confirm the function of SNIF-1 as IAA receptor we used a misexpression strategy. While AWA and AWC neurons induce attraction in worms, AWB neurons elicit avoidance response. (Troemel et al., 1997b, Bargmann, 2006, Bargmann et al., 1993, Troemel et al., 1997a). We reasoned that the misexpression of SNIF-1 in AWB should cause aversion to the odor. We generated transgenic worms carrying extrachromosomal array expressing snif-1 cDNA under a bonafide AWB-specific promoter (AWBp:: snif-1). As expected, the misexpression of snif-1 in AWB neurons elicited repulsion to IAA in worms (Figure 6H). To confirm the expression of snif-1 in AWC neurons, we generated a transgenic line of worms expressing GFP under snif-1 promoter, and mCherry under odr-1 promoter (to mark AWC neurons). We found that snif-1 is expressed faintly in many neurons, with strong expression in one of the two AWC neurons marked by odr-1p::mCherry (Figure 6I). Taken together, our findings show that C. elegans senses foraging signal, IAA, using SNIF-1 GPCR in one of the AWC neurons.

Based on our study, we propose a model describing an odor-dependent mechanism for sensing EAA-enriched bacteria in C. elegans. We show that IAA is an olfactory signal for leucine-enriched diet. AWC olfactory neurons of C. elegans allow worms to choose for IAA producing bacteria. Bacteria preferred by worms have the metabolic capacity to produce IAA from leucine via the Ehrlich degradation pathway. SNIF-1, a GPCR in AWC neurons, mediates foraging for bacteria in C. elegans by facilitating chemoperception of IAA (Figure 7).

Model depicting odor-based foraging strategy used by C. elegans C. elegans prefers leucine-enriched diets in an odor-dependent manner.

Preferred bacteria catabolize leucine, an EAA, to produce IAA via the Ehrlich degradation pathway. SRD-12/SNIF-1 GPCR expressed in AWC odor sensory neurons mediates the foraging behavior of C. elegans by sensing IAA.

Discussion

Our study provides evidence for an odor-dependent mechanism for foraging in C. elegans. Using CeMbio, a model microbiome for C. elegans, we show that worms prefer leucine-supplemented bacteria in an odor-dependent manner. Leucine-enriched diets produce significantly higher levels of IAA odor, making up to 90% of their headspace. These bacteria produce IAA from leucine using the Ehrlich degradation pathway. Preference for leucine-enriched bacteria solely depends on the AWC odor sensory neurons of C. elegans. Although the preferred bacteria produce several AWC-sensed odors, IAA exclusively influences the diet preference of worms. Phylogenetically distinct wild isolates of C. elegans, representing nine geographically distinct locations on Earth, respond robustly to IAA underscoring its relevance as a foraging signal in nature. C. elegans utilizes SNIF-1, a GPCR primarily expressed in AWC neurons, to sense IAA and forage for the preferred bacteria. Thus, IAA-SNIF-1 represents a ligand-receptor module used by C. elegans to forage for bacteria in its natural environment.

Is olfaction a major contributor to foraging behavior in animals in general? Artificial flowers, containing nectars with high amino acid content, are better at attracting butterflies (Alm et al., 1990). Honeybees prefer to consume artificial nectar rich in proline (Bertazzini et al., 2010). Hummingbirds, moths, and ants are attracted to flowers emitting benzyl acetone, an odor synthesized from an EAA, phenylalanine (Haverkamp et al., 2016, Yon et al., 2016, Bhattacharya and Baldwin, 2012, Kessler and Baldwin, 2007). For odors to serve as an attractant, the said odor must serve as a proxy or code for a needed primary metabolite, abundant in carbon, nitrogen, phosphorus, or micronutrients. In such a scenario, it is easy to envisage that each attractive cue might be a code for a specific essential nutrient.

Is isoamyl alcohol a signal for essential nutrient for C. elegans and other foraging animals. IAA was one of 61 attractive odors reported to be sensed at by C. elegans several decades ago (Bargmann et al., 1993), but its ecological relevance was unclear to date. A recent study also showed that worms specifically prefer IAA-producing bacteria (Chai et al., 2024, Worthy et al., 2018b). Our study suggests that IAA is the primary cue that drives foraging in favor of leucine-enriched diets in C. elegans. Does IAA reflect the metabolic state of bacteria producing it? IAA is indeed derived from a primary metabolite leucine, an essential branched chain amino acid for animals (Hazelwood et al., 2008). When bacteria were provided with additional leucine, the absolute abundance of IAA increases significantly in three different bacteria examined (Figures 2C and S3). The arrival of feeding insects to ripening fruits such as bananas coincides with for production of IAA and isoamyl acetate in the fruit via Ehrlich degradation pathway (Dou et al., 2022). The presence of Ehrlich degradation pathway in specific microbes and many insect attracting plants underscores the importance of odors in mediating interkingdom interactions. IAA, in combination with other odors, has been reported to attract Drosophila melanogaster and help in establishing a fly-fruit-yeast relationship (Becher et al., 2012, Dotterl and Gershenzon, 2023, He et al., 2022, Liu et al., 2022, Zjacic and Scholz, 2022, Haupt et al., 2010, Pham-Delegue et al., 1993). Why do animals need BCAA? Not only is leucine a major component of C. elegans proteome, intracellular accumulation of BCAA including leucine prolongs the lifespan of C. elegans. Edwards et al. 2015 have reported a 15% increase in the lifespan of worms upon 1 mM leucine supplementation (Edwards et al., 2015). Wang et al 2018 reported that along with lifespan extension, leucine supplementation makes worms more resistant to heat, paraquat, and UV-stress(Wang et al., 2018). The lack of response to leucine (Figure S1H in this study) in presence of IAA perception ability, via SNIF-1, suggests that worms rely on IAA sensing to identify bacteria with leucine. Additional natural microbiotas of C. elegans (not part of CeMbio) which produce IAA have also been reported to attract worms (Worthy et al., 2018a, Worthy et al., 2018c). Worms also prefer odors derived from other BCAA such as isobutanol derived from valine via Ehrlich degradation (Figure 4E) (Bizzio et al., 2021, Maoz et al., 2022, Yan et al., 2022). Phenylethyl alcohol (PEA in this study) can also be produced by Ehrlich degradation of phenyl alanine, another essential amino acid that animals require for their proteins as well as for the synthesis of neurotransmitter dopamine. All these reports support the idea that animals use olfactory cues to identify diets rich in essential amino acids. Animals use GPCRs to sense soluble and olfactory cues present in their environment. Rats utilize taste receptors T1R1/T1R3 for sensing alanine (Taylor-Burds et al., 2004).

C. elegans can also sense lysine and histidine (Ward, 1973) but not leucine (this study). Several studies have implicated olfactory GPCRs in regulating animal behavior. In termites, olfactory co-receptor OCRO influences foraging by enabling pheromone sensing (Xu et al., 2024). Or83b, an odorant receptor in D. melanogaster regulates lifespan (Libert et al., 2007). Six odor sensory neurons of C. elegans express hundreds of GPCRs, most of them remain orphan (Hammarlund et al., 2018).ODR-10 expressed in AWA odor sensory neurons regulates foraging behavior in C. elegans for lactic acid bacteria which produce diacetyl odor (Choi et al., 2016). AWC odor sensory neurons express as many as 181 GPCRs, of which 18 GPCRs display strong expression (Hammarlund et al., 2018) These receptors likely mediate the foraging behavior of worms. One of these, STR-2, senses 2-heptanone odor produced by Bacillus nematocidal (Zhang et al., 2016). STR-2 also regulates lipid accumulation and lifespan in a temperature-dependent manner (Dixit et al., 2020). The ligand(s) for the remaining 17 receptors were unknown. The discovery of one of these, SNIF-1/SRD-12, as receptor for sensing IAA (this study) suggests that the ligand for other receptors may also be found in the headspace of the C. elegans microbiome. Our study provides a framework for the identification of ligands for orphan GPCR, by searching for an ecologically relevant repertoire, the microbiome of an animal.

Methods

Experimental model and subject details Strains and growth media

C. elegans strains used in this study are listed in see Table S1. All strains were maintained as hermaphrodites at 20°C on nematode growth media (NGM) plates seeded with E. coli OP50, as originally described (Brenner, 1974). All the CeMbio strains used in this study (see Table S1) were grown on Difco Luria-Bertani (LB) media and maintained at 25°C, as described by Dirksen et al. (Dirksen et al., 2020). A list of all the reagents used in this study is provided in Table S1.

CRISPR approach for gene editing

Homology-directed integration of the ssDNA oligo was done in the srd-12 (snif-1) gene using a CRISPR/Cas9-based approach as described previously (Pu et al., 2023a). ssDNA oligo was designed to contain two stop codons along with a restriction site between two 35 bp long homology arms flanking the PAM sites of the targeted gene. ssDNA oligo was incubated with optimized RNP complexes of Cas9, crRNA, and tracrRNA. The mix was injected into the gonad of C. elegans along with rol-6 [pRF4::rol-6(su1006)] marker plasmid. The F1 progenies were segregated and genotyped using PCR, followed by restriction digestion for the introduction of mutation. srd-12 CRISPR mutants were back-crossed three times with N2 Bristol WT strain to generate VSL2401 and VSL2402.

Molecular Biology

The expression constructs used in this study were generated using the Multisite Gateway System (ThermoFisher Scientific). The promoters were amplified from genomic DNA and cloned in vector P4P1. For expressing srd-12(snif-1) under its native promoter, 2066 bp upstream region to CDS of srd-12 (snif-1) was used. The srd-12 (snif-1) encoding region was amplified using cDNA isolated from wild-type worms and cloned in pDONR221. LR reactions were performed to assemble the final construct. All constructs were verified by sequencing.

For generating transgenic lines, adult worms with a single row of eggs were injected with a mix consisting of cDNA-expressing constructs and co-marker at 50 ng/μl and 180 ng/μl, respectively. Coelomocyte marker (unc-122p::GFP) was used as a co-injection marker. 3-5 independent transgenic lines were generated before selecting a stable line.

Diet preference assay

For diet preference assays, individual CeMbio bacterium (test bacterium) and E. coli (control bacterium) were grown in LB broth overnight at 25°C and adjusted to an OD600 of 1. Standard diet choice assays were conducted on 90 mm NGM plates, wherein 25 µl of cultures of the test and control bacteria were spotted 1 cm away from the periphery of the plate on diametrically opposite points of the dish. These plates were incubated at 25°C for 12 hours. Gravid adult worms were washed thrice with S-basal buffer (100 mM NaCl, 5.75 mM K2HPO4, 44 mM KH2PO4). 50-70 worms were placed in the center of the assay plate and incubated at 25°C. The preference index (PI) was calculated after 3 hours using the following equation:

‘Odor-only’ diet preference assays were performed in tripartitioned plates, wherein one compartment was filled with buffered agar or BA (1 mM CaCl2, 1 mM MgSO4, 5 µg/ml cholesterol, 25 mM KPO4 buffer, and 2% agar) and other two compartments were filled with NGM. All the assay plates were air-dried for 70-90 minutes. For EAA supplementation, one of the NGM-filled compartments was supplemented with individual EAA at 5 mM concentration (+ EAA), while the other was left unsupplemented (-EAA). All bacterial strains were grown in LB broth overnight at 25°C and adjusted to an OD600 of 1. A volume of 25 µl of the test bacterial culture was spotted on the two compartments with NGM. The plates were incubated at 25°C for 12 hours. At the time of assay, 1 µl of sodium azide was spotted near the edges of the BA-containing compartment. Gravid adult worms were washed thrice with S-basal buffer, and 50-70 worms were placed at the center of the BA-containing compartment and incubated at 25°C (schematic in Figure 1A). The preference index (PI) was calculated after 3 hours using the following equation:

To understand worms’ preference for individual CeMbio bacterium (test bacterium) over E. coli OP50 (control bacterium), we performed an ‘odor-only’ diet preference assay as described above without any supplementations. The test bacterium culture was spotted in one of the NGM-filled compartments, while E. coli OP50 was spotted in the other compartment (schematic in Figure S2A). The remaining steps of the assay were performed in a manner similar to the one mentioned above. The PI was calculated using the following equation:

All diet choice experiments were conducted as standard preference assays unless otherwise specified.

Identification of volatiles produced by microbiota of C. elegans

The volatile profiles of CeMbio strains and E. coli OP50 were determined using gas chromatography-mass spectrometry (GC-MS/MS) (Prakash et al., 2021). Briefly, the individual bacterium was inoculated in 3 ml of LB broth and grown overnight at 25°C. NGM dishes (60 mm) were seeded with seven spots of 50 µl bacterial cultures adjusted to an OD600 of 1. Seeded plates were then incubated at 25°C for 23 hours. Two plates of the same conditions were sealed together using parafilm and incubated at 25°C for 1 hour to build odor concentration in the headspace for sampling. The volatiles were collected using a solid-phase microextraction fiber (SPME DVB: Divinylbenzene – Carboxen-WR – polydimethyl siloxane, 80 µm: Agilent Technologies, Part no. 5191-5874). Odors were subjected to thermal desorption and were identified using gas chromatography-mass spectrometry in an 8890C gas chromatography machine coupled with a 7000D GC/TQ, which used a capillary column HP-5MS ultra inert (30 m x 0.25 mm and 0.25 m, Agilent 19091S-433UI: 0245625H) with Helium as the carrier gas at a constant flow rate of 1.5 ml/min. The injection of the sample was done at 40°C. The GC program included a 40°C hold for 1 minute followed by a temperature ramp to 170°C at a rate of 5°C/min, then by a ramp to 270°C at a rate of 100°C/min, and finally, hold for 2 minutes. The inlet, M.S. source, and M.S. quadrupole temperatures were maintained at 225°C, 230°C, and 150°C, respectively. Odors were then identified using NIST (National Institute of Standards and Technology) 2019 V2.3 Mass Spectral Library and Agilent MassHunter Workstation version 10.0. The m/z peaks detected in the odor profile of NGM plates seeded with LB broth were subtracted from the odor profile of each tested bacterium. For leucine supplementation, preferred diets (CEent1, JUb66, and BIGb0170) were seeded on NGM supplemented with 5 mM leucine (+ LEU) and without supplementation (-LEU). GC-MS/MS was performed as described above, and a heat map was generated by analyzing the area under the curve of individual volatile peaks (Figure 2C).

For quantification of IAA, a standard curve was prepared by measuring the area under the curve for five concentrations (0.18 µM, 0.36 µM, 2.25 µM, 4.50 µM, and 8.99 µM) of IAA (Figure S3E). The preferred diets were seeded with two spots of 50 µl culture on NGM dishes supplemented with (+ LEU) and without (-LEU) leucine. The amount of IAA produced by each bacterium was determined by comparing the area under the curve in the GC-MS/MS plot of each bacterium against the standard curve for IAA.

Bacterial RNA extraction and qRT-PCR relative expression analysis

For RNA isolation, bacterial culture was grown in 20 ml of NGM broth with 5 mM leucine (+ LEU) and without leucine (-LEU) for 24 hours at 25°C. Bacterial cells were harvested by centrifugation at 5000 rpm for 20 min. The bacterial pellet was treated with RNAprotect Bacteria Reagent (Qiagen, Cat. No. 76506). Total RNA was extracted from the samples using the RNeasy Mini kit (Qiagen, Cat. No. 74104), and DNase I (NEB Cat. No. M0303S) treatment was done to remove genomic DNA. cDNA was synthesized using the iScript cDNA synthesis kit (Bio-Rad, Cat. No. 170-8891) and used for qRT-PCR to analyze the relative gene expression of ilvE using SYBR Green detection mix (Bio-Rad Cat. No. 1725124) on StepOnePlus (Applied Biosystems) machine. rboB was used as a housekeeping gene. The comparative ΔΔCt method was used to determine the fold change of the ilvE target gene.

Chemotaxis assay

Chemotaxis assays were performed as described previously (Prakash et al., 2021). Briefly, BA plates (90 mm) air-dried for 90 minutes were used for the assays. Sodium azide (2 µl) was spotted on the two diametrically opposite ends of the plate, followed by 2 µl of the test chemical to one side and solvent to the opposite side (schematic in Figure S4A). Test chemicals were diluted using suitable solvents (Table S1). Gravid adult worms were washed thrice with S-basal buffer. 50 to 80 gravid adult worms were placed in the center of the plate and incubated at 25°C for 3 hours. The chemotaxis index (CI) was calculated using the following equation:

For leucine, we performed modified chemotaxis assay adapted as described(Shingai et al., 2005, Ward, 1973). Briefly, 2 µl of leucine solution was spotted on one side of the assay plate and water on the opposite side. The plates were incubated at room temperature for 5 hours to create a leucine gradient. At the time of assay, sodium azide (2 µl) was spotted on the two diametrically opposite ends of the plate. The worms were prepared and introduced on the assay plate in a manner similar to the one mentioned above.

Odor adaptation

Gravid adult worms were washed thrice with S-basal buffer and transferred to 60 mm NGM plates. For odor adaptation, worms were exposed to bacterial odors by sealing an NGM plate seeded with bacteria (7 spots of 50 µL bacterial culture) with the plate containing washed worms. These plates were incubated at 25°C for 90 minutes to desensitize the worms to the odor. After exposure, worms were washed once with S basal buffer and used to perform chemotaxis assays (schematic in Figure 4A). For the modified odor adaptation assays, the NGM plate containing worms was exposed to IAA (1:10 dilution) by sealing it with another NGM agar plate containing 4 spots of 3 µl IAA or without IAA (schematic in Figure S6A). The adapted worms were then used for diet preference or chemotaxis assays as required.

Worm movement track analysis

We recorded the motion of five worms on a chemotaxis assay plate using a simple imaging setup. We used 1800 frames (1 fps) to track the behavior of worms. Using MATLAB, we extracted the set of binary images from the recorded frames and processed them using the Trackmate plugin of Fiji software to track worms’ motion. We used the Advanced Kalman algorithm to create tracks and interpolate for missing frames. The trajectory information was exported in the motilitylab spreadsheet format. This data was analyzed using MATLAB to measure the average speed of worms and the total distance traveled by worms (Saxton and Jacobson, 1997, Codling et al., 2008, Li et al., 2008, Berg, 1993).

Microscopy imaging

A synchronised population of adult worms were used for imaging. Animals were placed in 1 mM Sodium Azide on 2% agarose pads and imaged at 40X magnification. Imaging was done using a Zeiss LSM880 confocal microscope.

Statistical analyses

All statistical analyses were done using GraphPad Prism version 8. Statistical analyses were performed either by a two-tailed unpaired t-test between two groups or one-way ANOVA, followed by post hoc Dunnett’s multiple comparison test across multiple groups. The significant differences were denoted according to P-values: *P ≤ 0.1; **P ≤ 0.01; ***P ≤ 0.001; and ****P ≤ 0.0001. Data are presented as means ± SEM. All experiments were performed with 15 biological replicates done over three days.